---
output: html_document
title: "Real-data analysis: outputs for the GxPA LC-MS manuscript"
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "output/manuscript/", dpi = 300,
                      cache.path = "cache/manuscript/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "cowplot", "patchwork", "ggrepel"),
  library, character.only = TRUE))
select <- dplyr::select
theme_set(theme_bw())

dir.create("output/manuscript", recursive = TRUE, showWarnings = FALSE)
```

```{r load-results}
# Real-data model results are written by 02b_analysis.ipynb and copied down from
# the Terra workspace bucket into results/; the simulation writes its own tables
# to results/simulation/ (loaded further down, in the simulation section). Every section below degrades gracefully:
# if an input has not been copied yet, that section reports it and is skipped
# rather than failing the render.

# Results may sit inside the repo (results/) or at the project level (../results/,
# alongside doc/). Use whichever exists.
results_dir <- c("results", "../results") %>%
  keep(dir.exists) %>%
  pluck(1, .default = "results")

read_result <- function(f, dir = results_dir) {
  path <- file.path(dir, f)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE, progress = FALSE) %>%
    # The notebook writes a literal "β_GxM"/"β_G"/"β_E" column; normalise to ASCII.
    rename_with(~ gsub("β", "beta", .x))
}

have <- function(x) !is.null(x) && nrow(x) > 0

signif_gxm <- read_result("mesa_signif_GxMs.csv")   # Significant GxM hits, annotated
gxm_all    <- read_result("GxM_results.csv")        # Full metabolome-wide screen
y_g        <- read_result("Y_G.csv")                # Marginal SNP-outcome
e_g        <- read_result("E_G.csv")                # SNP-exposure (scenario-B diagnostic)
gxe        <- read_result("GxE.csv")                # The anchoring interactions
m_e        <- read_result("M_E.csv")                # Metabolite-exposure coupling
gxc        <- read_result("GxMpGxEpGxC.csv")        # G x covariate sensitivity
met_info   <- read_result("met_info-mesa.csv")      # Optional: metadata for all metabolites

# met_info-mesa.csv is the raw metadata written by 01a and carries no display
# label, so build one here rather than depending on the notebook to add it:
# annotated name where available, else m/z @ RT (Amide-neg has no RT).
if (have(met_info) && !("met_label" %in% names(met_info))) {
  met_info <- met_info %>%
    mutate(met_label = case_when(
      !is.na(Metabolite) ~ Metabolite,
      !is.na(RT)         ~ paste0("m/z ", round(MZ, 4), " @ ", round(RT, 2), " min"),
      TRUE               ~ paste0("m/z ", round(MZ, 4))))
}

# Ground truth for this table is the workspace bucket: on Terra only *.ipynb
# reaches the VM working directory, so the notebooks cannot read a repo-side CSV
# and 01b/02b/03c all read the bucket copy. Read the same copy they do, so this
# report and 02b can never disagree about gene assignments. Fall back to the repo
# copy when the bucket is unreachable, and say which was used.
voi_bucket <- paste0(Sys.getenv("WORKSPACE_BUCKET",
                                unset = "gs://fc-secure-4a392455-5587-4d6f-b8bd-01a1f834ae63"),
                     "/variants_of_interest.csv")
voi_tmp <- file.path(tempdir(), "variants_of_interest.csv")
voi_ok  <- system(paste("gcloud storage cp", shQuote(voi_bucket), shQuote(voi_tmp)),
                  ignore.stdout = TRUE, ignore.stderr = TRUE) == 0

variants_of_interest <- if (voi_ok) {
  read_csv(voi_tmp, show_col_types = FALSE)
} else if (file.exists("variants_of_interest.csv")) {
  read_csv("variants_of_interest.csv", show_col_types = FALSE)
} else NULL

voi_source <- if (voi_ok) {
  "workspace bucket (what the notebooks read)"
} else if (!is.null(variants_of_interest)) {
  "local repo copy - the bucket was unreachable"
} else {
  "not available"
}

# An edit committed to the repo but never published is invisible to every
# notebook, which is exactly how the two copies drifted before. Surface it.
if (voi_ok && file.exists("variants_of_interest.csv") &&
    !identical(readLines("variants_of_interest.csv", warn = FALSE),
               readLines(voi_tmp, warn = FALSE))) {
  voi_source <- paste(voi_source, "** REPO COPY HAS DRIFTED - publish it, see README **")
}

inputs <- tibble(
  File = c("mesa_signif_GxMs.csv", "GxM_results.csv", "Y_G.csv", "E_G.csv",
           "GxE.csv", "M_E.csv", "GxMpGxEpGxC.csv", "met_info-mesa.csv"),
  Contents = c("Significant GxM hits (annotated)", "Full metabolome-wide screen",
               "Marginal SNP-outcome", "SNP-exposure", "Anchoring GxE interactions",
               "Metabolite-exposure coupling", "G x covariate sensitivity",
               "Metabolite metadata (optional)"),
  Loaded = c(have(signif_gxm), have(gxm_all), have(y_g), have(e_g),
             have(gxe), have(m_e), have(gxc), have(met_info)),
  Rows = c(nrow(signif_gxm) %||% 0, nrow(gxm_all) %||% 0, nrow(y_g) %||% 0,
           nrow(e_g) %||% 0, nrow(gxe) %||% 0, nrow(m_e) %||% 0,
           nrow(gxc) %||% 0, nrow(met_info) %||% 0)
)

# Simulation inputs live in results/simulation/ and are listed separately, since
# 04_simulation.ipynb runs independently of the real-data pipeline and is often
# run on a different machine.
sim_files <- c("sim_params.csv"             = "Simulation parameters",
               "power_summary.csv"          = "Power, Scenario A",
               "power_z_replicates.csv"     = "Paired GxM z-statistics",
               "medmod_summary.csv"         = "Mediated-moderation comparison",
               "icc_grid_summary.csv"       = "ICC_E x ICC_M sensitivity grid",
               "tie_summary.csv"            = "Type I error, scenarios B/C/D",
               "tie_scenarioB_summary.csv"  = "Scenario B variants",
               "tie_scenarioC_summary.csv"  = "Scenario C variants",
               "tie_gamma_summary.csv"      = "Type I error, gamma-distributed E")

inputs <- bind_rows(inputs, tibble(
  File     = names(sim_files),
  Contents = unname(sim_files),
  Loaded   = file.exists(file.path(results_dir, "simulation", names(sim_files))),
  Rows     = NA_integer_
))
```

```{r load-qc}
# QC summaries written into results/qc/ by 01a, 01b, 01c, 02a and 02b.
#
# They exist because this report cannot read what they describe: the metabolite
# matrices and analysis_df-mesa.csv are individual-level TOPMed data and stay on
# Terra -- which is exactly why the redundancy section further down goes blank on a
# local knit. Every upstream decision that has to be auditable here therefore
# travels as counts, quantiles and per-feature statistics instead.
qc_files <- c(
  "sample_flow-mesa.csv"                  = "Sample flow, MESA exams to analysis rows",
  "metabolomics_lane_coverage-mesa.csv"   = "Coverage by LC-MS method",
  "imputation_ledger-mesa.csv"            = "Imputation counts, per variable per exam",
  "variable_distributions-mesa.csv"       = "Exposure/outcome distributions by exam",
  "lipid_medication-mesa.csv"             = "Lipid-lowering medication",
  "icc-mesa.csv"                          = "Within-person ICC, phenotypes",
  "metabolite_icc-mesa.csv"               = "Within-person ICC, per metabolite",
  "metabolite_icc_summary-mesa.csv"       = "Within-person ICC, per LC-MS method",
  "genotype_by_race-mesa.csv"             = "Allele frequency by self-identified race",
  "genotype_qc-mesa.csv"                  = "Call rate and dosage polarity",
  "table_s1_characteristics-mesa.csv"     = "Table S1, participant characteristics",
  "analysis_n_by_exam-mesa.csv"           = "Analysis n by exam",
  "covariate_ladder-mesa.csv"             = "Covariate ladder estimates",
  "covariate_variance_explained-mesa.csv" = "Variance in Y explained by each covariate set",
  "mpc_covariate_pvalues-mesa.csv"        = "Metabolite PC ~ covariate associations",
  "metabolite_qc_steps-mesa.csv"          = "Features and samples surviving each QC step",
  "metabolite_distributions-mesa.csv"     = "Missingness and QC-pool CV, binned",
  "metabolite_qc_quantiles-mesa.csv"      = "Missingness and QC-pool CV, quantiles",
  "m_eff_estimators-mesa.csv"             = "Effective number of tests, several estimators",
  "metabolite_eigenvalues-mesa.csv"       = "Metabolite eigenvalue spectrum")

qc <- list()
# NB single brackets and list(): `qc[[n]] <- NULL` would DELETE the element rather
# than record the absent file, and the inventory below would then be one row short
# of the files it claims to cover.
for (f in names(qc_files))
  qc[sub("-mesa\\.csv$|\\.csv$", "", f)] <- list(read_result(file.path("qc", f)))

inputs <- bind_rows(inputs, tibble(
  File     = file.path("qc", names(qc_files)),
  Contents = unname(qc_files),
  Loaded   = map_lgl(qc, have),
  Rows     = map_int(qc, ~ nrow(.x) %||% 0L)))

# One place to say "this section's input has not been copied down yet", so a
# partially populated results/qc/ reads as a gap rather than as a silence.
skip_note <- function(...) cat("Not available (", paste(c(...), collapse = ", "),
                              ") - skipping. Re-run the notebook that writes it, then ",
                              "copy results/qc/ down from the workspace bucket.\n", sep = "")
```

```{r params}
# Effective number of independent metabolites and the resulting screen threshold.
# Read from the notebook's own analysis_params.csv when available, so the two
# cannot drift apart; otherwise fall back to the value 02b last reported.
params <- read_result("analysis_params.csv")
N_EFF_METABOLITES <- if (have(params)) params$n_eff_metabolites[1] else 68.7
p_thresh_mwas     <- if (have(params)) params$p_thresh_mwas[1]     else 0.05 / N_EFF_METABOLITES
threshold_source  <- if (have(params)) "read from analysis_params.csv" else "fallback default"

# Readable labels for outcomes and exposures.
outcome_labels  <- c(hdl_log = "log(HDL-C)", tg_log = "log(TG)",
                     fg = "Fasting glucose", hb = "HbA1c")
exposure_labels <- c(mvpa_wins = "MVPA", bmi = "BMI", sex = "Sex", smoking = "Smoking")

relabel <- function(x, lookup) coalesce(unname(lookup[x]), x)

# Display names for the QC tables. The raw name is the auditable key, so it stays
# in the exported CSVs; only what a reader sees is relabelled.
qc_var_labels <- c(
  age = "Age (years)", bmi = "BMI (kg/m2)", hdl = "HDL-C (mg/dL)",
  hdl_log = "log(HDL-C)", tg = "Triglycerides (mg/dL)", tg_log = "log(TG)",
  fg = "Fasting glucose (mg/dL)", hb = "HbA1c (%)",
  mvpa = "MVPA (MET-min/week)", mvpa_wins = "MVPA, winsorized",
  mvpa_log = "log(MVPA + 1)", mvpa_trunc = "MVPA, truncated",
  pa = "Total PA", mod_pa = "Moderate PA", vig_pa = "Vigorous PA",
  ahei_score = "AHEI-2010 score", dash_score = "DASH score",
  ses_score = "Neighbourhood SES score", drinks_per_week = "Alcohol (drinks/week)",
  sex = "Sex", race = "Race/ethnicity (self-identified)", smoking = "Smoking",
  site = "Study site", income = "Income", season = "Season", month = "Month",
  insulin = "Insulin")

qc_label <- function(x) relabel(x, qc_var_labels)

qc_quantity_labels <- c(feature_missingness = "Missingness per feature",
                        sample_missingness  = "Missingness per sample",
                        qc_pool_cv          = "QC-pool CV")

# Anchor = one (outcome, exposure, SNP) combination taken into the screen.
# NB: the argument is `dat`, not `df` -- the results tables carry a column named
# `df` (Satterthwaite degrees of freedom), which would shadow the argument inside
# mutate()'s data mask and silently break the gene lookup.
anchor_label <- function(dat) {
  has_gene <- "gene" %in% names(dat)
  dat %>%
    mutate(anchor = paste0(if (has_gene) coalesce(gene, rsid) else rsid,
                           " (", rsid, ")\n",
                           relabel(E, exposure_labels), " -> ",
                           relabel(Y, outcome_labels)))
}
```

```{r anchor-scope}
# ---------------------------------------------------------------------------
# Pre-specified anchor set.
#
# 02b screens a superset of (SNP, outcome, exposure) combinations. The anchors
# carried into the manuscript are chosen on prior grounds -- a published
# interaction we are deliberately following up -- not on how they rank in MESA,
# and they are declared in anchors.csv so the choice lives in one auditable
# place rather than being implied by a filter somewhere downstream.
#
# To change the reported scope, edit anchors.csv: flip `include` to FALSE to
# park an anchor (its rows stay in results/, they just stop being reported).
# Dropping the CETP locus to a physical-activity-only paper is one edit there.
# ---------------------------------------------------------------------------

anchors <- if (file.exists("anchors.csv")) {
  read_csv("anchors.csv", show_col_types = FALSE) %>% filter(include)
} else {
  NULL
}

if (is.null(anchors)) {
  warning("anchors.csv not found - reporting every anchor present in the results.")
  in_scope    <- function(dat, ...) dat
  scope_note  <- "no anchors.csv found; every screened anchor is reported"
} else {
  # Anchor-keyed tables (GxM, GxE, GxCovar) match on all three of G, Y, E.
  in_scope <- function(dat) {
    if (is.null(dat) || nrow(dat) == 0) return(dat)
    keys <- intersect(c("G", "Y", "E"), names(dat))
    if (length(keys) == 0) return(dat)
    semi_join(dat, distinct(anchors[keys]), by = keys)
  }
  scope_note <- paste(anchors$label, collapse = "; ")
}

# Marginal tables are keyed by only part of the anchor, so they get their own
# filters rather than being forced through in_scope().
keep_pairs <- function(dat, a, b) {
  if (is.null(dat) || nrow(dat) == 0 || is.null(anchors)) return(dat)
  semi_join(dat, distinct(anchors[c(a, b)]), by = c(a, b))
}

gxm_all    <- in_scope(gxm_all)
gxe        <- in_scope(gxe)
gxc        <- in_scope(gxc)
signif_gxm <- in_scope(signif_gxm)      # hit selection is per-row, so filtering
                                        # and re-selecting give the same rows.
y_g        <- keep_pairs(y_g, "Y", "G")
e_g        <- keep_pairs(e_g, "E", "G")
m_e        <- if (have(signif_gxm) && have(m_e))
                semi_join(m_e, distinct(select(signif_gxm, M, E)), by = c("M", "E")) else m_e
```


# Input inventory

```{r input-inventory}
inputs %>%
  mutate(Loaded = ifelse(Loaded, "yes", "MISSING")) %>%
  kable(caption = "Result files available to this report") %>%
  kable_styling(full_width = FALSE)
```

Variant table read from: **`r voi_source`**

Anchors reported: **`r scope_note`**
(pre-specified in `anchors.csv`; anchors with `include = FALSE` remain in
`results/` but are not reported here).

Screen-wide significance threshold: **p < `r signif(p_thresh_mwas, 3)`**
(Bonferroni over `r signif(N_EFF_METABOLITES, 4)` effective metabolites;
`r threshold_source`). Note that this corrects across metabolites within an
anchor, not across anchors.

# Upstream QC

Everything in this section is rendered from `results/qc/`, written by the
preprocessing notebooks. Nothing here is recomputed, and no individual-level data
is read -- the point is that the decisions made before any model was fitted are
visible in the same document as the models, rather than only in the notebook that
made them.

## Sample construction

```{r qc-sample-flow}
if (have(qc$sample_flow)) {
  flow_tbl <- qc$sample_flow %>%
    filter(exam == "all") %>%
    arrange(step) %>%
    select(Step = step, Stage = stage, Rows = n_rows, Participants = n_individuals,
           `Rows lost` = rows_dropped, `Participants lost` = individuals_dropped,
           Reason = reason)

  write_csv(flow_tbl, "output/manuscript/qc_sample_flow.csv")

  flow_tbl %>%
    kable(caption = paste("Sample flow. Filters are cumulative and in the order the",
                          "analysis applies them; a row is one participant-exam",
                          "observation. The last row is the n the primary model sees.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/sample_flow-mesa.csv")
```

```{r qc-sample-flow-by-exam}
if (have(qc$sample_flow)) {
  qc$sample_flow %>%
    filter(exam != "all") %>%
    arrange(step) %>%
    select(Stage = stage, exam, n_rows) %>%
    pivot_wider(names_from = exam, names_prefix = "Exam ", values_from = n_rows) %>%
    kable(caption = paste("Sample flow by exam. Exam 4 carries no physical-activity",
                          "questionnaire, and metabolomics was run on a subset of exams,",
                          "so the losses are not spread evenly.")) %>%
    kable_styling(full_width = FALSE)
}
```

```{r qc-lane-coverage}
if (have(qc$metabolomics_lane_coverage)) {
  qc$metabolomics_lane_coverage %>%
    filter(exam == "all") %>%
    select(`LC-MS method` = lane, Features = n_features,
           Observations = n_rows, Participants = n_individuals) %>%
    kable(caption = paste("Coverage by LC-MS method. A sample can pass QC on one",
                          "method and fail on another, so the n behind a given",
                          "metabolite depends on which method it came from.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/metabolomics_lane_coverage-mesa.csv")
```

## Imputation

```{r qc-imputation}
if (have(qc$imputation_ledger)) {
  imp_tbl <- qc$imputation_ledger %>%
    filter(exam == "all", n_missing_before > 0) %>%
    arrange(desc(n_missing_before)) %>%
    select(Variable = variable, Imputation = imputation, Rows = n_rows,
           Missing = n_missing_before, `% missing` = frac_missing_before,
           Imputed = n_imputed, `Still missing` = n_still_missing) %>%
    mutate(`% missing` = signif(100 * `% missing`, 3))

  write_csv(imp_tbl, "output/manuscript/qc_imputation_ledger.csv")

  imp_tbl %>%
    kable(caption = paste("Imputation ledger, pooled over exams. Counts come from the",
                          "`__imp` flags 01c sets before any filling, so they record what",
                          "was actually missing. Every imputed value is one the models",
                          "treat as observed.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/imputation_ledger-mesa.csv")
```

```{r qc-imputation-by-exam, fig.asp = 0.5}
if (have(qc$imputation_ledger)) {
  qc$imputation_ledger %>%
    filter(exam != "all") %>%
    ggplot(aes(x = exam, y = reorder(qc_label(variable), n_missing_before),
               fill = 100 * frac_missing_before)) +
    geom_tile() +
    scale_fill_viridis_c(name = "% missing\nbefore imputation", option = "magma",
                         direction = -1) +
    labs(x = "MESA exam", y = NULL)
}
```

## Exposure, outcome and covariate distributions

```{r qc-mvpa-transforms}
if (have(qc$variable_distributions)) {
  qc$variable_distributions %>%
    filter(grepl("^mvpa", variable)) %>%
    mutate(variable = qc_label(variable)) %>%
    select(Variable = variable, Exam = exam, n, `% zero` = n_zero, Mean = mean,
           SD = sd, Min = min, P25 = p25, Median = median, P75 = p75, Max = max) %>%
    mutate(`% zero` = signif(100 * `% zero` / n, 3),
           across(c(Mean, SD, Min, P25, Median, P75, Max), ~ signif(.x, 3))) %>%
    kable(caption = paste("MVPA under each transformation 02a compared. `mvpa_wins` is",
                          "the one the models use: winsorized at the 90th percentile,",
                          "which is what moves the maximum.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/variable_distributions-mesa.csv")
```

```{r qc-distributions-by-exam, fig.asp = 0.55}
if (have(qc$variable_distributions)) {
  qc$variable_distributions %>%
    filter(exam != "all",
           variable %in% c("hdl", "hdl_log", "tg", "bmi", "age", "mvpa", "mvpa_wins", "fg")) %>%
    mutate(exam = factor(exam), variable = qc_label(variable)) %>%
    ggplot(aes(x = exam, y = median)) +
    geom_pointrange(aes(ymin = p25, ymax = p75), size = 0.3) +
    geom_linerange(aes(ymin = p05, ymax = p95), alpha = 0.35) +
    facet_wrap(vars(variable), scales = "free_y") +
    labs(x = "MESA exam", y = "Median (P25-P75, P05-P95)",
         caption = paste("A step in HDL across exams would be assay or secular drift",
                         "rather than biology; MVPA is absent at exam 4."))
}
```

Lipid-lowering medication is reported here for context, **not** as an adjustment,
and HDL-C is deliberately left uncorrected. The convention (GLGC and essentially
every large lipid GWAS since) corrects TC and LDL-C for treatment and leaves HDL-C
and TG alone: the statin effect on LDL-C is large and consistent enough to justify a
single divisor, while its effect on HDL-C is a few percent and varies by drug and
dose, so a correction factor would add more error than it removes. Adjusting for
medication as a *covariate* would be worse still -- treatment is prescribed in
response to lipid levels, so it is a descendant of the outcome, and conditioning on
it risks collider bias in exactly the E -> Y slope the interaction models are about.

The table is useful in two other ways: as the denominator for a sensitivity analysis
refitting the hits with treated participants excluded, and as a flag on the
metabolite side, where statins do have a detectable signature and a GxM hit on a
statin-responsive feature would be worth a second look.

```{r qc-lipid-meds}
if (have(qc$lipid_medication)) {
  qc$lipid_medication %>%
    mutate(`%` = signif(100 * frac, 3)) %>%
    select(Variable = variable, Value = value, Participants = n_participants, `%`) %>%
    kable(caption = paste("Lipid-lowering medication prevalence. Reported for context;",
                          "no lipid trait in this pipeline is medication-corrected.")) %>%
    kable_styling(full_width = FALSE)
} else {
  cat("No lipid-medication summary in results/qc/. 01c writes an empty table when the",
      "PIC-SURE pull in 00 returned no such column - add one there if the report",
      "should carry this.\n")
}
```

## Within-person reliability (ICC)

The simulation assumes ICC = 0.5 for both the exposure and the mediator. These are
the same quantity measured in MESA, which is what connects the two halves of the
paper: if the real ICCs sit well below the assumed value, the simulation's power
curves are optimistic for this dataset.

Read them as a **lower bound** on reliability. The exams are 5-10 years apart, so
the within-person term mixes measurement error with genuine change, and anything
that really moves over a decade -- MVPA above all -- is penalised for moving.

```{r qc-icc, fig.asp = 0.5}
if (have(qc$icc) || have(qc$metabolite_icc_summary)) {
  # One frame for both halves of the panel: phenotypes are single estimates,
  # metabolites are a distribution over features, and they share the y axis.
  icc_panel <- bind_rows(
    if (have(qc$icc))
      qc$icc %>% arrange(desc(icc)) %>%
        transmute(label = qc_label(variable), kind = "Phenotype", point = icc) else NULL,
    if (have(qc$metabolite_icc_summary))
      qc$metabolite_icc_summary %>%
        transmute(label = gsub("_", "-", lane), kind = "Metabolites (per feature)",
                  p05, p25, median, p75, p95) else NULL) %>%
    mutate(label = fct_inorder(label),
           kind  = factor(kind, levels = c("Phenotype", "Metabolites (per feature)")))

  ggplot(icc_panel, aes(x = label)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "firebrick") +
    { if ("median" %in% names(icc_panel))
        geom_boxplot(aes(ymin = p05, lower = p25, middle = median, upper = p75,
                         ymax = p95),
                     stat = "identity", width = 0.55, fill = "grey85", na.rm = TRUE) } +
    { if ("point" %in% names(icc_panel))
        geom_point(aes(y = point), size = 2, na.rm = TRUE) } +
    facet_grid(cols = vars(kind), scales = "free_x", space = "free_x") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = "ICC across exams",
         caption = paste("Dashed line: the ICC = 0.5 the simulation assumes.",
                         "Boxes are the P25-P75 of the per-feature ICCs, whiskers P05-P95.")) +
    theme(axis.text.x = element_text(angle = 35, hjust = 0.9))
} else skip_note("qc/icc-mesa.csv", "qc/metabolite_icc_summary-mesa.csv")
```

```{r qc-icc-table}
if (have(qc$icc)) {
  icc_tbl <- qc$icc %>%
    arrange(desc(icc)) %>%
    mutate(variable = qc_label(variable)) %>%
    select(Variable = variable, Participants = n_individuals,
           Observations = n_observations, `Between-person var` = var_between,
           `Within-person var` = var_within, ICC = icc) %>%
    mutate(across(where(is.numeric) & !c(Participants, Observations), ~ signif(.x, 3)))

  write_csv(icc_tbl, "output/manuscript/qc_icc.csv")

  icc_tbl %>%
    kable(caption = paste("Within-person ICC(1) across exams, among participants with",
                          "at least two measured values. Carried-forward values are",
                          "excluded: they are copies of the previous exam's and would",
                          "drive the ICC towards 1.")) %>%
    kable_styling(full_width = FALSE)
}
```

```{r qc-icc-hits}
# The hits' own reliability is the part of this that bears on the results: a hit on
# a feature with near-zero ICC is a hit on something the assay barely reproduces.
if (have(qc$metabolite_icc) && have(signif_gxm)) {
  hit_icc <- signif_gxm %>%
    select(M, met_label) %>%
    distinct() %>%
    inner_join(select(qc$metabolite_icc, M = feature, icc, n_individuals), by = "M")

  if (nrow(hit_icc)) {
    hit_icc %>%
      arrange(desc(icc)) %>%
      select(Metabolite = met_label, Participants = n_individuals, ICC = icc) %>%
      mutate(ICC = signif(ICC, 3)) %>%
      kable(caption = "Within-person ICC of the significant metabolites") %>%
      kable_styling(full_width = FALSE)
  } else cat("No overlap between the hits and the metabolites with repeated measures.\n")
}
```

## Covariate justification

```{r qc-covariate-ladder, fig.asp = 0.5}
if (have(qc$covariate_ladder)) {
  qc$covariate_ladder %>%
    filter(outcome == "hdl_log", !grepl("^chr", variable)) %>%
    mutate(covar_set = fct_inorder(covar_set), variable = qc_label(variable)) %>%
    ggplot(aes(x = covar_set, y = estimate, color = covar_set)) +
    geom_hline(yintercept = 0, color = "gray70") +
    geom_pointrange(aes(ymin = l95, ymax = u95), size = 0.3) +
    facet_wrap(vars(variable), scales = "free_y") +
    scale_x_discrete(labels = NULL, breaks = NULL) +
    labs(x = NULL, y = "Standardised estimate",
         color = "Covariate set",
         caption = "Rungs are cumulative; the last is the set 02b fits.")
} else skip_note("qc/covariate_ladder-mesa.csv")
```

```{r qc-variance-explained}
if (have(qc$covariate_variance_explained)) {
  ve_tbl <- qc$covariate_variance_explained %>%
    mutate(outcome = qc_label(outcome),
           across(c(r2, adj_r2, delta_adj_r2), ~ signif(.x, 3))) %>%
    select(Outcome = outcome, `Covariate set` = covar_set, n, Terms = n_terms,
           `R2` = r2, `Adj. R2` = adj_r2, `Delta adj. R2` = delta_adj_r2)

  write_csv(ve_tbl, "output/manuscript/qc_covariate_variance_explained.csv")

  ve_tbl %>%
    kable(caption = paste("Variance explained by each covariate set. All sets are fitted",
                          "on the same complete-case rows, so the rungs are comparable;",
                          "the delta column is the gain from adding that rung.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/covariate_variance_explained-mesa.csv")
```

Each cell below is one `mPC ~ covariate` linear model. Only exam-1 samples with
data on all four LC-MS methods contribute to the metabolite PCs, so this is a
statement about that subset rather than about the whole analysis sample.

```{r qc-mpc-heatmap, fig.asp = 0.6}
if (have(qc$mpc_covariate_pvalues)) {
  qc$mpc_covariate_pvalues %>%
    mutate(mPC = factor(mPC, levels = paste0("mPC", 1:100)),
           nlp = pmin(-log10(p), 10)) %>%
    ggplot(aes(x = mPC, y = fct_rev(fct_inorder(covariate)), fill = nlp)) +
    geom_tile() +
    geom_text(data = ~ filter(.x, p < 0.05), label = "*", color = "white", size = 2.5,
              vjust = 0.75) +
    scale_fill_gradient(name = expression(-log[10](italic(p))),
                        low = "black", high = "steelblue", limits = c(0, 10)) +
    labs(x = NULL, y = NULL, caption = "* marks p < 0.05. Shading capped at 10.") +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9))
} else skip_note("qc/mpc_covariate_pvalues-mesa.csv")
```

## Metabolite QC

```{r qc-met-steps}
if (have(qc$metabolite_qc_steps)) {
  steps_tbl <- qc$metabolite_qc_steps %>%
    # Step index and label are separate columns: a label starting "1." is read as
    # an ordered-list marker inside a table cell and swallowed by pandoc.
    { if ("step_label" %in% names(.)) . else mutate(., step_label = as.character(step)) } %>%
    select(Method = method_label, Step = step, Stage = step_label,
           Features = n_features, Samples = n_samples,
           `Features dropped` = features_dropped, `Samples dropped` = samples_dropped)

  write_csv(steps_tbl, "output/manuscript/qc_metabolite_steps.csv")

  steps_tbl %>%
    kable(caption = paste("Features and samples surviving each 01a QC step. The last",
                          "row is the merged matrix: a sample profiled by only some",
                          "methods still contributes, which is why it has more samples",
                          "than any single method.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/metabolite_qc_steps-mesa.csv")
```

```{r qc-met-distributions, fig.asp = 0.4}
if (have(qc$metabolite_distributions)) {
  qc$metabolite_distributions %>%
    # The three quantities use different breaks, so the bins are ordered by their
    # lower edge rather than by order of appearance in the file.
    mutate(quantity = factor(recode(quantity, !!!qc_quantity_labels),
                             levels = unname(qc_quantity_labels)),
           bin = fct_reorder(bin, bin_lower)) %>%
    ggplot(aes(x = bin, y = n, fill = method_label)) +
    geom_col(position = "dodge") +
    facet_wrap(vars(quantity), scales = "free") +
    labs(x = NULL, y = "Count", fill = "Method") +
    theme(axis.text.x = element_text(angle = 40, hjust = 0.9))
} else skip_note("qc/metabolite_distributions-mesa.csv")
```

```{r qc-met-quantiles}
if (have(qc$metabolite_qc_quantiles)) {
  qc$metabolite_qc_quantiles %>%
    mutate(quantity = recode(quantity, !!!qc_quantity_labels),
           across(c(mean, p05, p25, median, p75, p95), ~ signif(.x, 3))) %>%
    select(Method = method_label, Quantity = quantity, n, Mean = mean,
           P05 = p05, P25 = p25, Median = median, P75 = p75, P95 = p95,
           any_of(c(`QC-pool samples` = "n_qc_pool_samples"))) %>%
    kable(caption = paste("Missingness and pooled-QC CV. The CV is taken on the raw",
                          "intensity scale before the log2, over the pooled QC samples",
                          "01a drops, so it measures analytical reproducibility rather",
                          "than biological spread.")) %>%
    kable_styling(full_width = FALSE)
}
```

### The screen threshold

```{r qc-meff}
if (have(qc$m_eff_estimators)) {
  meff_tbl <- qc$m_eff_estimators %>%
    mutate(Used = ifelse(in_use, "yes", ""),
           m_eff = signif(m_eff, 4), p_threshold = signif(p_threshold, 3)) %>%
    select(Estimator = estimator, `M_eff` = m_eff, `p threshold` = p_threshold, Used)

  write_csv(meff_tbl, "output/manuscript/qc_m_eff.csv")

  meff_tbl %>%
    kable(caption = paste("Effective number of independent metabolites under several",
                          "estimators, over", qc$m_eff_estimators$n_features[1],
                          "features in", qc$m_eff_estimators$n_samples[1],
                          "samples with complete data on all four methods. The choice",
                          "of estimator moves the threshold by more than an order of",
                          "magnitude, which is why it is reported here rather than",
                          "asserted.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/m_eff_estimators-mesa.csv")
```

```{r qc-eigenvalues, fig.asp = 0.42}
if (have(qc$metabolite_eigenvalues)) {
  spec <- qc$metabolite_eigenvalues

  p_spec <- spec %>%
    filter(component <= 200) %>%
    ggplot(aes(x = component, y = eigenvalue)) +
    geom_line() +
    scale_y_log10() +
    labs(x = "Component", y = "Eigenvalue (log scale)")

  p_cum <- spec %>%
    ggplot(aes(x = component, y = cum_var_explained)) +
    geom_line() +
    geom_hline(yintercept = c(0.95, 0.99), linetype = "dashed", color = "gray50") +
    { if (have(qc$m_eff_estimators))
        geom_vline(data = filter(qc$m_eff_estimators, in_use),
                   aes(xintercept = m_eff), color = "firebrick") } +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Component", y = "Cumulative variance explained",
         caption = "Red line: the M_eff the screen threshold uses.")

  p_spec + p_cum
} else skip_note("qc/metabolite_eigenvalues-mesa.csv")
```

## Genotype QC

```{r qc-genotype}
if (have(qc$genotype_qc)) {
  geno_tbl <- qc$genotype_qc %>%
    { if (!is.null(anchors)) filter(., cpaid %in% anchors$G | rsid %in% anchors$rsid) else . } %>%
    mutate(across(c(alt_freq, maf, call_rate), ~ signif(.x, 3))) %>%
    select(any_of(c("gene", "rsid", "cpaid")), `Counted allele` = counted_allele,
           `Freq (counted)` = alt_freq, MAF = maf, Called = n_called,
           `Call rate` = call_rate)

  write_csv(geno_tbl, "output/manuscript/qc_genotype.csv")

  geno_tbl %>%
    kable(caption = paste("Call rate and dosage polarity for the reported anchors.",
                          "01b re-polarizes a flipped variant so every dosage counts the",
                          "allele named in its `cpaid`; a variant whose counted allele is",
                          "undefined has no named alt in variants_of_interest.csv, and",
                          "the sign of its effect is relative to an unlabelled allele.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/genotype_qc-mesa.csv")
```

```{r qc-genotype-by-race}
if (have(qc$genotype_by_race)) {
  qc$genotype_by_race %>%
    { if (!is.null(anchors)) filter(., variant %in% anchors$G) else . } %>%
    mutate(across(c(freq_counted, maf, call_rate), ~ signif(.x, 3))) %>%
    select(Variant = variant, Race = race, Participants = n_participants,
           Called = n_called, `Freq (counted)` = freq_counted, MAF = maf,
           `Call rate` = call_rate) %>%
    kable(caption = paste("Allele frequency by self-identified race. MESA is",
                          "multi-ethnic, so a frequency that swings across groups is",
                          "the reason the models adjust for genetic PCs and fit",
                          "gPC x E interaction terms.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/genotype_by_race-mesa.csv")
```

## Table S1. Participant characteristics

```{r qc-table-s1}
if (have(qc$table_s1_characteristics)) {
  s1 <- qc$table_s1_characteristics %>%
    mutate(cell = case_when(
      variable == "N"   ~ as.character(n),
      !is.na(level)     ~ paste0(n, " (", sprintf("%.1f", pct), "%)"),
      TRUE              ~ paste0(signif(mean, 3), " ± ", signif(sd, 2))),
      row = ifelse(is.na(level), qc_label(variable),
                   paste0(qc_label(variable), ": ", level))) %>%
    select(row, group, cell) %>%
    pivot_wider(names_from = group, values_from = cell) %>%
    # A cell is empty only when that level does not occur in that stratum.
    mutate(across(-row, ~ replace_na(.x, "-")))

  write_csv(s1, "output/manuscript/table_s1_characteristics.csv")

  s1 %>%
    rename(` ` = row) %>%
    kable(caption = paste("Table S1. Characteristics of the primary analysis sample",
                          "(exam-1 rows with metabolomics, genotype at the anchor SNP",
                          "and HDL-C measured). Continuous variables are mean +/- SD,",
                          "categorical are n (%).")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/table_s1_characteristics-mesa.csv")
```

```{r qc-n-by-exam}
if (have(qc$analysis_n_by_exam)) {
  qc$analysis_n_by_exam %>%
    select(Exam = exam, Observations = n_rows, Participants = n_individuals) %>%
    kable(caption = paste("Observations contributed by each exam under the same",
                          "filters. The longitudinal models pool these, which is why",
                          "the model n exceeds the number of participants.")) %>%
    kable_styling(full_width = FALSE)
}
```

# Stage 0: how do the anchoring interactions look in MESA?

The anchors are chosen on prior grounds, so these tables are reported for
transparency rather than as a gate: they show how strongly each published
interaction is expressed in MESA under the current model specification, before
any metabolite is examined. A weak anchor is informative in its own right --
the simulation predicts that the GxM test is least specific exactly there.

```{r anchor-gxe}
if (have(gxe)) {
  gxe_tbl <- gxe %>%
    filter(!is.na(est)) %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene, analysis),
                  by = "G") else . } %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    arrange(p) %>%
    select(any_of(c("Outcome", "Exposure", "rsid", "gene", "analysis")),
           beta_GxE = est, se, p, n) %>%
    mutate(across(c(beta_GxE, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(gxe_tbl, "output/manuscript/anchor_gxe.csv")

  gxe_tbl %>%
    kable(caption = "Anchoring GxE interactions in MESA") %>%
    kable_styling(full_width = FALSE)
} else cat("GxE.csv not available - skipping.\n")
```

```{r anchor-marginal}
if (have(y_g) || have(e_g)) {
  marg <- bind_rows(
    if (have(y_g)) mutate(filter(y_g, !is.na(est)), Model = "Y ~ G",
                          Trait = relabel(Y, outcome_labels)) else NULL,
    if (have(e_g)) mutate(filter(e_g, !is.na(est)), Model = "E ~ G",
                          Trait = relabel(E, exposure_labels)) else NULL
  ) %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    arrange(p) %>%
    select(Model, Trait, any_of(c("rsid", "gene")), beta = est, se, p, n) %>%
    mutate(across(c(beta, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(marg, "output/manuscript/anchor_marginal.csv")

  marg %>%
    head(30) %>%
    kable(caption = paste("Marginal SNP associations (top 30). E ~ G is the",
                          "diagnostic for an upstream (scenario B) architecture:",
                          "a strong G -> E effect would suggest the SNP acts on",
                          "the exposure rather than downstream of it.")) %>%
    kable_styling(full_width = FALSE)
} else cat("Y_G.csv / E_G.csv not available - skipping.\n")
```

# Metabolome-wide interaction screen

```{r mwis-counts}
if (have(gxm_all)) {
  gxm_annot <- gxm_all %>%
    filter(!is.na(est)) %>%
    { if (have(met_info))
        left_join(., select(met_info, M = unique_met_id, Metabolite, met_label,
                            Method, MZ, RT, HMDB_ID),
                  by = "M") else . } %>%
    { if (!("met_label" %in% names(.)) && have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    mutate(hit = p < p_thresh_mwas)

  screen_summary <- gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    group_by(across(any_of(c("rsid", "gene"))), Y, E) %>%
    summarise(`Metabolites tested` = n(),
              `Hits` = sum(hit),
              `Min p` = signif(min(p, na.rm = TRUE), 3),
              .groups = "drop") %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    select(any_of(c("rsid", "gene")), Outcome, Exposure,
           `Metabolites tested`, Hits, `Min p`) %>%
    arrange(desc(Hits))

  write_csv(screen_summary, "output/manuscript/screen_summary.csv")

  screen_summary %>%
    kable(caption = "Metabolome-wide screen: hits per anchor") %>%
    kable_styling(full_width = FALSE)
} else cat("GxM_results.csv not available - skipping.\n")
```

```{r mwis-volcano, fig.asp = 0.5}
if (have(gxm_all)) {
  volcano_df <- gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    anchor_label()

  volcano_plt <- volcano_df %>%
    ggplot(aes(x = est, y = -log10(p))) +
    geom_point(aes(color = hit), size = 0.7, alpha = 0.6) +
    geom_hline(yintercept = -log10(p_thresh_mwas), linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, color = "gray80") +
    { if ("met_label" %in% names(volcano_df))
        geom_text_repel(data = filter(volcano_df, hit), aes(label = met_label),
                        size = 2.2, min.segment.length = 0, box.padding = 0.6,
                        point.padding = 0.2, segment.size = 0.3, segment.color = "gray50",
                        max.overlaps = Inf, seed = 1, direction = "y", hjust = 0) } +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.28))) +
    scale_color_manual(values = c(`FALSE` = "gray70", `TRUE` = "firebrick"),
                       guide = "none") +
    facet_wrap(vars(anchor), scales = "free_x") +
    labs(x = expression("G x M interaction estimate (" * beta[GxM] * ")"),
         y = expression(-log[10](italic(p))))

  volcano_plt
} else cat("GxM_results.csv not available - skipping.\n")
```

```{r mwis-by-method, fig.asp = 0.45}
if (have(gxm_all) && "Method" %in% names(gxm_annot)) {
  gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    anchor_label() %>%
    arrange(Method) %>%
    group_by(anchor) %>%
    mutate(idx = row_number()) %>%
    ungroup() %>%
    ggplot(aes(x = idx, y = -log10(p), color = Method)) +
    geom_point(size = 0.6, alpha = 0.7) +
    geom_hline(yintercept = -log10(p_thresh_mwas), linetype = "dashed", color = "gray40") +
    facet_wrap(vars(anchor)) +
    labs(x = "Metabolite (ordered by LC-MS method)",
         y = expression(-log[10](italic(p)))) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
} else cat("Metabolite metadata (Method) not available - skipping per-method plot.\n")
```

```{r signif-hits-table}
if (have(signif_gxm)) {
  hits_tbl <- signif_gxm %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    arrange(p) %>%
    select(any_of(c("rsid", "gene")), Outcome, Exposure,
           Metabolite = met_label, any_of(c("HMDB_ID", "Method", "MZ", "RT")),
           beta_GxM, se, p) %>%
    mutate(across(any_of(c("MZ")), ~ signif(.x, 6)),
           across(any_of(c("RT")), ~ signif(.x, 3)),
           across(c(beta_GxM, se, p), ~ signif(.x, 3)))

  write_csv(hits_tbl, "output/manuscript/table1_gxm_hits.csv")

  hits_tbl %>%
    kable(caption = "Table 1. Significant gene x metabolite interactions") %>%
    kable_styling(full_width = FALSE)
} else cat("mesa_signif_GxMs.csv not available - skipping.\n")
```

# Directness: how much interaction signal does GxM recover that GxE does not?

The conceptual model says the observed GxE interaction is a *diluted* version of
the GxM interaction: with E and M standardised and coupled by
`M = alpha*E + eps`, we expect `beta_GxE ~= alpha * beta_GxM`. Dilution has two
sources -- imperfect measurement of E, and the structural fact that M has
determinants other than E -- and both push `alpha` below 1 regardless of how well
E is measured. The tables below quantify that gap directly in the MESA data.

```{r gxm-vs-gxe}
if (have(signif_gxm) && have(gxe)) {
  gxe_key <- gxe %>%
    filter(!is.na(est)) %>%
    select(Y, E, G, beta_GxE = est, se_GxE = se, p_GxE = p)

  directness <- signif_gxm %>%
    inner_join(gxe_key, by = c("Y", "E", "G")) %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels),
           ratio_beta = beta_GxM / beta_GxE,
           gain_nlp = -log10(p) - (-log10(p_GxE))) %>%
    arrange(p) %>%
    select(any_of(c("rsid", "gene")), Outcome, Exposure, Metabolite = met_label,
           beta_GxM, p_GxM = p, beta_GxE, p_GxE,
           `beta ratio (GxM/GxE)` = ratio_beta,
           `-log10p gain` = gain_nlp) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(directness, "output/manuscript/directness_gxm_vs_gxe.csv")

  directness %>%
    kable(caption = paste("GxM versus the anchoring GxE. A beta ratio > 1 and a",
                          "positive -log10p gain indicate the metabolite carries",
                          "the interaction more directly than the exposure does.")) %>%
    kable_styling(full_width = FALSE)
} else cat("Need both mesa_signif_GxMs.csv and GxE.csv - skipping.\n")
```

```{r dilution-check, fig.asp = 0.5}
# Quantitative test of the conceptual model: predicted beta_GxE = alpha * beta_GxM,
# where alpha is the standardised M ~ E coefficient. Deviations are informative --
# they bound how much of the GxE signal the exposure can carry at all.
if (have(signif_gxm) && have(gxe) && have(m_e)) {
  alpha_df <- m_e %>%
    filter(!is.na(est)) %>%
    select(M, E, alpha = est, alpha_se = se, alpha_p = p) %>%
    distinct(M, E, .keep_all = TRUE)  # One alpha per (metabolite, exposure); guards the join below.

  dilution <- signif_gxm %>%
    inner_join(select(filter(gxe, !is.na(est)), Y, E, G, beta_GxE = est),
               by = c("Y", "E", "G")) %>%
    inner_join(alpha_df, by = c("M", "E")) %>%
    mutate(predicted_GxE = alpha * beta_GxM,
           Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels))

  if (nrow(dilution) > 0) {
    dil_tbl <- dilution %>%
      arrange(p) %>%
      select(any_of(c("rsid", "gene")), Outcome, Exposure, Metabolite = met_label,
             `alpha (M~E)` = alpha, beta_GxM, `predicted beta_GxE` = predicted_GxE,
             `observed beta_GxE` = beta_GxE) %>%
      mutate(across(where(is.numeric), ~ signif(.x, 3)))

    write_csv(dil_tbl, "output/manuscript/dilution_check.csv")

    print(kable_styling(kable(
      dil_tbl,
      caption = paste("Dilution check: is the observed GxE consistent with",
                      "alpha * beta_GxM? alpha is the standardised metabolite-exposure",
                      "coupling, so a small alpha implies the exposure can only ever",
                      "carry a fraction of the interaction."))))

    dilution %>%
      ggplot(aes(x = predicted_GxE, y = beta_GxE)) +
      geom_abline(slope = 1, intercept = 0, color = "gray60", linetype = "dashed") +
      geom_hline(yintercept = 0, color = "gray85") +
      geom_vline(xintercept = 0, color = "gray85") +
      geom_point(aes(color = Exposure), size = 2) +
      geom_text_repel(aes(label = met_label), size = 2.5, max.overlaps = 15) +
      labs(x = expression("Predicted " * beta[GxE] * " ( = " * alpha %*% beta[GxM] * " )"),
           y = expression("Observed " * beta[GxE]))
  } else cat("No overlapping (M, E) rows between the hits and M_E.csv - skipping.\n")
} else cat("Need mesa_signif_GxMs.csv, GxE.csv and M_E.csv - skipping.\n")
```

# Are the hits plausible mediators?

A metabolite that mediates the exposure's contribution should itself be
associated with the exposure. Metabolites with a strong GxM but no M~E
association are candidates for the reverse-causation or confounding
architectures rather than true mediation.

```{r m-e-table}
if (have(m_e)) {
  me_tbl <- m_e %>%
    filter(!is.na(est)) %>%
    { if (have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    { if (have(met_info) && !("met_label" %in% names(.)))
        left_join(., select(met_info, M = unique_met_id, met_label), by = "M") else . } %>%
    mutate(Exposure = relabel(E, exposure_labels)) %>%
    arrange(p) %>%
    select(Metabolite = any_of(c("met_label", "M")), Exposure,
           beta_E = est, se, p, n) %>%
    mutate(across(c(beta_E, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(me_tbl, "output/manuscript/metabolite_exposure.csv")

  me_tbl %>%
    kable(caption = "Metabolite-exposure associations for the significant hits") %>%
    kable_styling(full_width = FALSE)
} else cat("M_E.csv not available - skipping.\n")
```

# Redundancy among the hits

Significant metabolites are often strongly correlated with one another, so a raw
count of hits overstates the number of independent signals. This section
collapses them where the QC'd metabolite matrix is available locally.

```{r hit-correlations, fig.asp = 0.8}
met_matrix_path <- "data/derived/metabolomics/QCd/merged_QCd-mesa.csv"

if (have(signif_gxm) && file.exists(met_matrix_path)) {
  hit_ids <- unique(signif_gxm$M)
  met_mat <- read_csv(met_matrix_path, show_col_types = FALSE, progress = FALSE) %>%
    select(any_of(hit_ids))

  if (ncol(met_mat) > 1) {
    cmat <- cor(met_mat, use = "pairwise.complete.obs")
    lab <- signif_gxm$met_label[match(colnames(cmat), signif_gxm$M)]
    dimnames(cmat) <- list(lab, lab)

    clust <- hclust(as.dist(1 - abs(cmat)))
    n_indep <- length(unique(cutree(clust, h = 1 - 0.8)))  # Merge at |r| > 0.8

    cat("Independent signal clusters at |r| < 0.8:", n_indep,
        "from", ncol(cmat), "significant metabolites\n\n")

    as_tibble(cmat, rownames = "met1") %>%
      pivot_longer(-met1, names_to = "met2", values_to = "r") %>%
      mutate(across(c(met1, met2), ~ factor(.x, levels = clust$labels[clust$order]))) %>%
      ggplot(aes(x = met1, y = met2, fill = r)) +
      geom_tile() +
      scale_fill_gradient2(name = "Pearson r", limits = c(-1, 1)) +
      labs(x = "", y = "") +
      theme(axis.text.x = element_text(angle = 40, hjust = 0.9))
  } else cat("Fewer than two significant metabolites - nothing to cluster.\n")
} else {
  cat("Metabolite matrix not available locally (", met_matrix_path, ").\n",
      "Copy it down, or run this section on Terra, to collapse correlated hits.\n", sep = "")
}
```

# Sensitivity: G x covariate adjustment

```{r gxcovars}
if (have(gxc)) {
  gxc_tbl <- gxc %>%
    filter(!is.na(est)) %>%
    { if (have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    mutate(Outcome = relabel(Y, outcome_labels)) %>%
    arrange(p) %>%
    select(any_of(c("rsid", "gene")), Outcome,
           Metabolite = any_of(c("met_label", "M")),
           beta_GxM_adj = est, se, p) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(gxc_tbl, "output/manuscript/sensitivity_gxcovars.csv")

  gxc_tbl %>%
    kable(caption = paste("GxM estimates with additional G x covariate adjustment.",
                          "Estimates that attenuate substantially relative to Table 1",
                          "suggest the interaction is confounded rather than causal.")) %>%
    kable_styling(full_width = FALSE)
} else cat("GxMpGxEpGxC.csv not available - skipping.\n")
```

```{r attenuation-comparison}
if (have(gxc) && have(signif_gxm)) {
  atten <- signif_gxm %>%
    select(Y, E, G, M, met_label, beta_primary = beta_GxM, p_primary = p) %>%
    inner_join(select(filter(gxc, !is.na(est)), Y, E, G, M,
                      beta_adj = est, p_adj = p),
               by = c("Y", "E", "G", "M")) %>%
    mutate(`% attenuation` = 100 * (1 - beta_adj / beta_primary)) %>%
    arrange(p_primary) %>%
    select(Metabolite = met_label, beta_primary, beta_adj,
           `% attenuation`, p_primary, p_adj) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(atten, "output/manuscript/sensitivity_attenuation.csv")

  atten %>%
    kable(caption = "Primary versus G x covariate-adjusted GxM estimates") %>%
    kable_styling(full_width = FALSE)
}
```

# Simulation study

Operating characteristics of the screening pipeline, computed by
`04_simulation.ipynb` and read here from `results/simulation/`. The notebook
writes rejection rates and per-replicate p-values; every figure below is a
rendering of one of those tables, so a restyle or a re-facet costs a knit rather
than a re-run of the simulation.

```{r load-sim}
sim_dir <- file.path(results_dir, "simulation")

read_sim <- function(f) {
  path <- file.path(sim_dir, f)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE, progress = FALSE)
}

sim_params  <- read_sim("sim_params.csv")
sim_power   <- read_sim("power_summary.csv")
sim_power_z <- read_sim("power_z_replicates.csv")
sim_medmod  <- read_sim("medmod_summary.csv")
sim_icc     <- read_sim("icc_grid_summary.csv")
sim_tie     <- read_sim("tie_summary.csv")
sim_tie_b   <- read_sim("tie_scenarioB_summary.csv")
sim_tie_c   <- read_sim("tie_scenarioC_summary.csv")
sim_gamma   <- read_sim("tie_gamma_summary.csv")
dag_nodes   <- read_sim("dag_nodes.csv")
dag_edges   <- read_sim("dag_edges.csv")

sim_param <- function(name, default = NA) {
  if (!have(sim_params)) return(default)
  v <- sim_params$value[sim_params$param == name]
  if (length(v) == 0) default else v[1]
}
SIM_ALPHA <- as.numeric(sim_param("ALPHA", 0.05))

# The exported tables carry raw keys, so every display label lives here. That
# keeps one styling layer for the whole report rather than one in the notebook
# and another here, which is what let the two drift apart previously.
test_labels <- c(p_gxe        = "GxE",         p_gxm     = "GxM",
                 p_med        = "Mediation",
                 p_unadj      = "Unadjusted",  unadj     = "Unadjusted",
                 p_adj_GxE    = "G:E-adjusted", adjGxE   = "G:E-adjusted",
                 p_adj_U_main = "U-adjusted",   adjU     = "U-adjusted",
                 p_adj_GxU    = "G:U-adjusted", adjGxU   = "G:U-adjusted")
variant_labels <- c(full         = "Baseline",
                    full_adj_GxE = "+ G:E adjustment",
                    no_marg      = "No marginal E-M correlation")
icc_labels  <- c(perfect = "Perfect measurement",
                 default = "Realistic measurement error")
medmod_labels <- c(perfect = "Perfect (ICC = 1)", realistic = "Realistic (ICC = 0.5)")
scenario_labels <- c(A = "Molecular mediator (primary)", B = "Upstream GxE",
                     C = "Reverse causation",            D = "E-M Confounding")

sim_cols   <- c(GxE = "#d95f02", GxM = "#1b9e77", Mediation = "#7570b3")
sim_shapes <- c(GxE = 17, GxM = 16, Mediation = 15)
D_MODELS   <- c("Unadjusted", "U-adjusted", "G:U-adjusted", "G:E-adjusted")

# Rate-vs-parameter curves (power sweeps) and rate bar charts (type I error)
# are the only two figure forms in the simulation, so they get one definition
# each rather than one per panel.
sim_curve <- function(dat, xvar, xlab, ylab = paste0("Rejection rate at alpha = ", SIM_ALPHA)) {
  ggplot(dat, aes({{ xvar }}, rate, color = test, fill = test, shape = test)) +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey50") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2) +
    scale_color_manual(values = sim_cols) +
    scale_fill_manual(values = sim_cols) +
    scale_shape_manual(values = sim_shapes) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
    labs(x = xlab, y = ylab, color = NULL, fill = NULL, shape = NULL) +
    theme(legend.position = "bottom")
}

sim_bars <- function(dat, title, fill = "#1b9e77") {
  pd <- position_dodge(width = 0.7)
  ggplot(dat, aes(model, rate, alpha = condition)) +
    geom_col(fill = fill, color = "grey20", width = 0.7, position = pd) +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.18, position = pd) +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey40") +
    # Each panel uses a different fill, so the alpha key is drawn in a neutral
    # grey: identical guides across panels are what lets patchwork collect them
    # into the single shared legend below.
    scale_alpha_manual(values = c(0.45, 1.0),
                       guide = guide_legend(override.aes = list(fill = "grey30"))) +
    scale_x_discrete(labels = scales::label_wrap(14)) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(x = NULL, y = "Type I error rate of GxM test", alpha = NULL, title = title) +
    theme(axis.text.x = element_text(size = 8))
}

# Shared reshaping: map raw keys onto display labels and order the factors.
label_tests <- function(dat) mutate(dat, test = unname(test_labels[test]))
label_icc   <- function(dat) mutate(dat, condition = factor(unname(icc_labels[icc_condition]),
                                                            levels = unname(icc_labels)))
```

## Causal scenarios

```{r sim-dags, fig.width = 11, fig.asp = 0.30}
# The notebook writes the DAG geometry already resolved (edges trimmed back off
# the node labels, modifier arrows anchored to the midpoint of the edge they
# modify, coefficient labels offset perpendicular to their arrow), so this only
# has to draw it. Rendering here rather than in the notebook keeps every
# manuscript figure on one styling layer.
build_dag <- function(which_scenarios) {
  if (!have(dag_nodes) || !have(dag_edges)) return(NULL)
  e <- filter(dag_edges, scenario %in% which_scenarios)
  n <- filter(dag_nodes, scenario %in% which_scenarios)
  ggplot() +
    geom_segment(data = e,
                 aes(x = x, y = y, xend = xend, yend = yend,
                     linetype = kind, color = kind),
                 arrow = arrow(length = unit(0.10, "in"), type = "closed"),
                 linewidth = 0.55) +
    geom_text(data = e, aes(x = label_x, y = label_y, label = label, color = kind),
              parse = TRUE, size = 3.7, show.legend = FALSE) +
    geom_label(data = n, aes(x = x, y = y, label = node),
               size = 4.6, fontface = "bold", fill = "white",
               label.padding = unit(0.22, "lines")) +
    facet_wrap(~ scenario, nrow = 1, labeller = labeller(scenario = scenario_labels)) +
    scale_linetype_manual(values = c(main = "solid", mod = "dashed")) +
    scale_color_manual(values    = c(main = "grey20", mod = "firebrick")) +
    coord_cartesian(xlim = c(-0.2, 2.2), ylim = c(-0.4, 1.5), clip = "off", expand = FALSE) +
    theme_void(base_size = 13) +
    theme(legend.position = "none",
          strip.text    = element_text(face = "bold"),
          panel.spacing = unit(0.2, "lines"),
          plot.margin   = margin(2, 2, 2, 2))
}

dag_all <- build_dag(c("A", "B", "C", "D"))
if (!is.null(dag_all)) print(dag_all) else
  cat("results/simulation/dag_nodes.csv and dag_edges.csv not available - skipping.\n")
```

Solid arrows are main effects; dashed red arrows are interaction terms, drawn
landing on the edge they modify. Scenario A is the mediation model the screen
assumes; B, C and D are the non-mediating structures that can still produce a
GxM signal.

## Power under true mediation (Scenario A)

```{r sim-power, fig.asp = 0.5}
if (have(sim_power)) {
  power_panel <<- sim_curve(label_tests(sim_power), beta_GxM,
                            expression(paste("True ", beta[GxM])))
  print(power_panel)

  write_csv(sim_power, "output/manuscript/sim_power.csv")
} else cat("results/simulation/power_summary.csv not available - skipping.\n")
```

The `G:E`-adjusted GxM test is the remedy deployed against the spurious
scenarios below. Comparing the per-replicate GxM z-statistic with and without
that adjustment shows it is free under true mediation: the two statistics fall
on the identity line.

```{r sim-power-z, fig.asp = 0.6}
if (have(sim_power_z)) {
  r <- cor(sim_power_z$z_gxm, sim_power_z$z_gxm_adj)

  power_z_panel <<-
    ggplot(sim_power_z, aes(z_gxm, z_gxm_adj, color = factor(beta_GxM))) +
      geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
      geom_point(alpha = 0.5, size = 1.3) +
      scale_color_brewer(palette = "Greens", name = expression(paste("True ", beta[GxM]))) +
      coord_equal() +
      labs(x = "GxM z (unadjusted)", y = "GxM z (G:E-adjusted)",
           subtitle = sprintf("Pearson r = %.4f", r)) +
      theme(legend.position = "bottom")
  print(power_z_panel)
} else cat("results/simulation/power_z_replicates.csv not available - skipping.\n")
```

## Comparison with formal mediated-moderation testing

```{r sim-medmod, fig.asp = 0.5}
if (have(sim_medmod)) {
  dat <- sim_medmod %>%
    label_tests() %>%
    mutate(test      = factor(test, levels = c("GxE", "GxM", "Mediation")),
           condition = factor(unname(medmod_labels[condition]),
                              levels = unname(medmod_labels)))

  print(sim_curve(dat, beta_EM, expression(paste("E-M coupling ", beta[EM]))) +
          facet_wrap(~ condition))
} else cat("results/simulation/medmod_summary.csv not available - skipping.\n")
```

## Sensitivity: power across the full ICC_E x ICC_M grid

```{r sim-icc-grid, fig.asp = 0.95}
if (have(sim_icc)) {
  dat <- sim_icc %>%
    label_tests() %>%
    mutate(ICC_E_label = factor(sprintf("ICC[E] == %.2f", ICC_E)),
           ICC_M_label = factor(sprintf("ICC[M] == %.2f", ICC_M)))

  print(sim_curve(dat, beta_GxM, expression(paste("True ", beta[GxM]))) +
          facet_grid(ICC_E_label ~ ICC_M_label, labeller = label_parsed))
} else cat("results/simulation/icc_grid_summary.csv not available - skipping.\n")
```

## Type I error under the non-mediating scenarios

```{r sim-tie, fig.width = 11, fig.asp = 0.42}
tie_variant_panel <- function(dat, title, fill) {
  if (!have(dat)) return(NULL)
  dat %>%
    label_icc() %>%
    mutate(model = factor(unname(variant_labels[variant]),
                          levels = unname(variant_labels))) %>%
    sim_bars(title, fill)
}

panel_b <<- tie_variant_panel(sim_tie_b, "Upstream GxE", "#7570b3")
panel_c <<- tie_variant_panel(sim_tie_c, "Reverse causation", "#d95f02")

panel_d <<- if (have(sim_tie)) {
  sim_tie %>%
    filter(scenario == "D", test != "p_gxe") %>%
    label_icc() %>%
    label_tests() %>%
    mutate(model = factor(test, levels = D_MODELS)) %>%
    sim_bars("E-M confounder", "#1b9e77")
} else NULL

panels <- compact(list(panel_b, panel_c, panel_d))
if (length(panels)) {
  # Scenario D has four bars against three in B and C, so it gets the extra width.
  widths <- ifelse(map_lgl(panels, ~ identical(.x, panel_d)), 1.3, 1)
  print(wrap_plots(panels, nrow = 1) +
          plot_layout(guides = "collect", widths = widths) &
          theme(legend.position = "bottom"))
} else cat("No scenario B/C/D summaries available - skipping.\n")
```

```{r sim-tie-table}
if (have(sim_tie)) {
  sim_tie %>%
    filter(!is.na(rate)) %>%
    label_tests() %>%
    transmute(Scenario = scenario,
              Measurement = unname(icc_labels[icc_condition]),
              Model = test,
              `Rejection rate` = round(rate, 3),
              `95% CI` = sprintf("%.3f - %.3f", lo, hi),
              Replicates = n_used) %>%
    arrange(Scenario, Measurement, Model) %>%
    kable(caption = "Type I error of the GxM test by scenario and adjustment") %>%
    kable_styling(full_width = FALSE)

  write_csv(sim_tie, "output/manuscript/sim_type1_error.csv")
}
```

## Combined manuscript figure

```{r sim-combined, fig.width = 12, fig.asp = 0.92}
# Reassembles the panels built above; nothing is recomputed. Missing inputs drop
# their row rather than failing the figure.
top_row <- if (!is.null(build_dag("A")) && exists("power_panel") && exists("power_z_panel")) {
  build_dag("A") + power_panel + power_z_panel + plot_layout(widths = c(1, 1.4, 1.2))
} else NULL

mid_row <- build_dag(c("B", "C", "D"))

tie_panels <- compact(list(
  if (exists("panel_b")) panel_b, if (exists("panel_c")) panel_c,
  if (exists("panel_d")) panel_d))
bot_row <- if (length(tie_panels)) {
  wrap_plots(tie_panels, nrow = 1) + plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
} else NULL

rows <- compact(list(top_row, mid_row, bot_row))
if (length(rows) == 3) {
  combined_fig <- rows[[1]] / rows[[2]] / rows[[3]] +
    plot_layout(heights = c(1.0, 0.8, 1.3)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 16))
  print(combined_fig)
} else {
  cat("Need the DAG tables, the power summaries and the type I error summaries",
      "to assemble the combined figure -", length(rows), "of 3 rows available.\n")
}
```

## Sensitivity: nonnegative (gamma-distributed) exposure

Rejection rates under `E ~ Gamma(4, 2)` against the normally-distributed
exposure at perfect measurement, so the two columns differ only in the shape of
E.

```{r sim-gamma, fig.asp = 0.5}
if (have(sim_gamma) && have(sim_tie)) {
  normal_arm <- sim_tie %>%
    filter(icc_condition == "perfect") %>%
    transmute(scenario, model = unname(test_labels[test]),
              rate, e_dist = "E ~ N(0, 1)")

  gamma_arm <- sim_gamma %>%
    transmute(scenario, model = unname(test_labels[test]),
              rate, e_dist = "E ~ Gamma(4, 2)")

  bind_rows(normal_arm, gamma_arm) %>%
    filter(!is.na(rate), model %in% D_MODELS) %>%
    mutate(model    = factor(model, levels = D_MODELS),
           e_dist   = factor(e_dist, levels = c("E ~ N(0, 1)", "E ~ Gamma(4, 2)")),
           scenario = paste("Scenario", scenario)) %>%
    ggplot(aes(model, rate, fill = e_dist)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65, color = "grey20") +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey40") +
    facet_wrap(~ scenario, scales = "free_x") +
    scale_fill_manual(values = c("E ~ N(0, 1)" = "#7570b3", "E ~ Gamma(4, 2)" = "#d95f02")) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(x = NULL, y = "Type I error rate of GxM test", fill = "Exposure distribution") +
    theme(legend.position = "bottom", axis.text.x = element_text(size = 8, angle = 30, hjust = 1))
} else cat("Need tie_gamma_summary.csv and tie_summary.csv - skipping.\n")
```

# Session

```{r session-info}
sessionInfo()
```
