---
output: html_document
title: "Real-data analysis: outputs for the GxPA LC-MS manuscript"
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "output/manuscript/", dpi = 300,
                      cache.path = "cache/manuscript/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "cowplot", "patchwork", "ggrepel"),
  library, character.only = TRUE))
select <- dplyr::select
theme_set(theme_bw())

dir.create("output/manuscript", recursive = TRUE, showWarnings = FALSE)
```

```{r load-results}
# Real-data model results are written by 02b_analysis.ipynb and copied down from
# the Terra workspace bucket into results/; the simulation writes its own tables
# to results/simulation/ (loaded further down, in the simulation section). Every section below degrades gracefully:
# if an input has not been copied yet, that section reports it and is skipped
# rather than failing the render.

# Results may sit inside the repo (results/) or at the project level (../results/,
# alongside doc/). Use whichever exists.
results_dir <- c("results", "../results") %>%
  keep(dir.exists) %>%
  pluck(1, .default = "results")

read_result <- function(f, dir = results_dir) {
  path <- file.path(dir, f)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE, progress = FALSE) %>%
    # The notebook writes a literal "β_GxM"/"β_G"/"β_E" column; normalise to ASCII.
    rename_with(~ gsub("β", "beta", .x))
}

have <- function(x) !is.null(x) && nrow(x) > 0

signif_gxm <- read_result("mesa_signif_GxMs.csv")   # Significant GxM hits, annotated
gxm_all    <- read_result("GxM_results.csv")        # Full metabolome-wide screen
y_g        <- read_result("Y_G.csv")                # Marginal SNP-outcome
e_g        <- read_result("E_G.csv")                # SNP-exposure (scenario-B diagnostic)
gxe        <- read_result("GxE.csv")                # The anchoring interactions
m_e        <- read_result("M_E.csv")                # Metabolite-exposure coupling
gxc        <- read_result("GxMpGxEpGxC.csv")        # G x covariate sensitivity
met_info   <- read_result("met_info-mesa.csv")      # Optional: metadata for all metabolites

# met_info-mesa.csv is the raw metadata written by 01a and carries no display
# label, so build one here rather than depending on the notebook to add it:
# annotated name where available, else m/z @ RT (Amide-neg has no RT).
if (have(met_info) && !("met_label" %in% names(met_info))) {
  met_info <- met_info %>%
    mutate(met_label = case_when(
      !is.na(Metabolite) ~ Metabolite,
      !is.na(RT)         ~ paste0("m/z ", round(MZ, 4), " @ ", round(RT, 2), " min"),
      TRUE               ~ paste0("m/z ", round(MZ, 4))))
}

# Ground truth for the hand-maintained tables is the workspace bucket: on Terra only
# *.ipynb reaches the VM working directory, so the notebooks cannot read a repo-side CSV
# and read the bucket copies. Read the same copies they do, so this report and the
# notebooks can never disagree about gene assignments or about which lanes exist. Fall
# back to the repo copy when the bucket is unreachable, and say which was used.
config_bucket <- Sys.getenv("WORKSPACE_BUCKET",
                            unset = "gs://fc-secure-4a392455-5587-4d6f-b8bd-01a1f834ae63")
read_config <- function(name) {
  tmp <- file.path(tempdir(), name)
  ok  <- system(paste("gcloud storage cp", shQuote(paste0(config_bucket, "/", name)), shQuote(tmp)),
                ignore.stdout = TRUE, ignore.stderr = TRUE) == 0
  dat <- if (ok) {
    read_csv(tmp, show_col_types = FALSE)
  } else if (file.exists(name)) {
    read_csv(name, show_col_types = FALSE)
  } else NULL
  source <- if (ok) {
    "workspace bucket (what the notebooks read)"
  } else if (!is.null(dat)) {
    "local repo copy - the bucket was unreachable"
  } else {
    "not available"
  }
  # An edit committed to the repo but never published is invisible to every notebook.
  if (ok && file.exists(name) &&
      !identical(readLines(name, warn = FALSE), readLines(tmp, warn = FALSE)))
    source <- paste(source, "** REPO COPY HAS DRIFTED - publish it, see README **")
  list(data = dat, source = source)
}

voi_cfg <- read_config("variants_of_interest.csv")
variants_of_interest <- voi_cfg$data
voi_source <- voi_cfg$source

inputs <- tibble(
  File = c("mesa_signif_GxMs.csv", "GxM_results.csv", "Y_G.csv", "E_G.csv",
           "GxE.csv", "M_E.csv", "GxMpGxEpGxC.csv", "met_info-mesa.csv"),
  Contents = c("Significant GxM hits (annotated)", "Full metabolome-wide screen",
               "Marginal SNP-outcome", "SNP-exposure", "Anchoring GxE interactions",
               "Metabolite-exposure coupling", "G x covariate sensitivity",
               "Metabolite metadata (optional)"),
  Loaded = c(have(signif_gxm), have(gxm_all), have(y_g), have(e_g),
             have(gxe), have(m_e), have(gxc), have(met_info)),
  Rows = c(nrow(signif_gxm) %||% 0, nrow(gxm_all) %||% 0, nrow(y_g) %||% 0,
           nrow(e_g) %||% 0, nrow(gxe) %||% 0, nrow(m_e) %||% 0,
           nrow(gxc) %||% 0, nrow(met_info) %||% 0)
)

# Simulation inputs live in results/simulation/ and are listed separately, since
# 04_simulation.ipynb runs independently of the real-data pipeline and is often
# run on a different machine.
sim_files <- c("sim_params.csv"             = "Simulation parameters",
               "power_summary.csv"          = "Power, Scenario A",
               "power_bias_summary.csv"     = "Bias and coverage of the GxM estimate",
               "power_z_replicates.csv"     = "Paired GxM z-statistics",
               "power_z_summary.csv"        = "GxM z agreement within effect size",
               "dilution_replicates.csv"    = "Dilution identity, per replicate",
               "dilution_summary.csv"       = "Dilution identity, summary",
               "medmod_summary.csv"         = "Mediated-moderation comparison",
               "icc_grid_summary.csv"       = "ICC_E x ICC_M sensitivity grid",
               "tie_summary.csv"            = "Type I error, scenarios B/C/D",
               "tie_scenarioB_summary.csv"  = "Scenario B variants",
               "tie_scenarioC_summary.csv"  = "Scenario C variants",
               "tie_scenarioB_nsweep_summary.csv" = "Scenario B over sample size",
               "tie_gamma_summary.csv"      = "Type I error, gamma-distributed E")

inputs <- bind_rows(inputs, tibble(
  File     = names(sim_files),
  Contents = unname(sim_files),
  Loaded   = file.exists(file.path(results_dir, "simulation", names(sim_files))),
  Rows     = NA_integer_
))
```

```{r load-qc}
# QC summaries written into results/qc/ by 01a, 01b, 01c, 02a and 02b.
#
# They exist because this report cannot read what they describe: the metabolite
# matrices and analysis_df-mesa.csv are individual-level TOPMed data and stay on
# Terra -- which is exactly why the redundancy section further down goes blank on a
# local knit. Every upstream decision that has to be auditable here therefore
# travels as counts, quantiles and per-feature statistics instead.
qc_files <- c(
  "sample_flow-mesa.csv"                  = "Sample flow, MESA exams to analysis rows",
  "metabolomics_method_coverage-mesa.csv" = "Coverage by LC-MS method",
  "imputation_ledger-mesa.csv"            = "Imputation counts, per variable per exam",
  "variable_distributions-mesa.csv"       = "Exposure/outcome distributions by exam",
  "icc-mesa.csv"                          = "Within-person ICC, phenotypes",
  "metabolite_icc-mesa.csv"               = "Within-person ICC, per metabolite",
  "metabolite_icc_summary-mesa.csv"       = "Within-person ICC, per LC-MS method",
  "genotype_by_race-mesa.csv"             = "Allele frequency by self-identified race",
  "genotype_qc-mesa.csv"                  = "Call rate and dosage polarity",
  "table_s1_characteristics-mesa.csv"     = "Table S1, participant characteristics",
  "analysis_n_by_exam-mesa.csv"           = "Analysis n by exam",
  "covariate_ladder-mesa.csv"             = "Covariate ladder estimates",
  "covariate_variance_explained-mesa.csv" = "Variance in Y explained by each covariate set",
  "mpc_covariate_pvalues-mesa.csv"        = "Metabolite PC ~ covariate associations",
  "metabolite_qc_steps-mesa.csv"          = "Features and samples surviving each QC step",
  "metabolite_distributions-mesa.csv"     = "Missingness and QC-pool CV, binned",
  "metabolite_qc_quantiles-mesa.csv"      = "Missingness and QC-pool CV, quantiles",
  "m_eff_estimators-mesa.csv"             = "Effective number of tests, several estimators",
  "metabolite_eigenvalues-mesa.csv"       = "Metabolite eigenvalue spectrum")

qc <- list()
# NB single brackets and list(): `qc[[n]] <- NULL` would DELETE the element rather
# than record the absent file, and the inventory below would then be one row short
# of the files it claims to cover.
for (f in names(qc_files))
  qc[sub("-mesa\\.csv$|\\.csv$", "", f)] <- list(read_result(file.path("qc", f)))

# 01c called LC-MS methods "lanes" until lane came to mean a declared (G, Y, E) analysis.
# Accept files written under the old names, so a results/qc/ that predates the rename
# still renders.
if (!have(qc$metabolomics_method_coverage)) {
  legacy <- read_result("qc/metabolomics_lane_coverage-mesa.csv")
  if (have(legacy)) qc["metabolomics_method_coverage"] <- list(rename(legacy, method = lane))
}
for (n in c("metabolite_icc", "metabolite_icc_summary"))
  if (have(qc[[n]]) && "lane" %in% names(qc[[n]]) && !("method" %in% names(qc[[n]])))
    qc[n] <- list(rename(qc[[n]], method = lane))

inputs <- bind_rows(inputs, tibble(
  File     = file.path("qc", names(qc_files)),
  Contents = unname(qc_files),
  Loaded   = map_lgl(qc, have),
  Rows     = map_int(qc, ~ nrow(.x) %||% 0L)))

# One place to say "this section's input has not been copied down yet", so a
# partially populated results/qc/ reads as a gap rather than as a silence.
skip_note <- function(...) cat("Not available (", paste(c(...), collapse = ", "),
                              ") - skipping. Re-run the notebook that writes it, then ",
                              "copy results/qc/ down from the workspace bucket.\n", sep = "")
```

```{r params}
# Effective number of independent metabolites and the resulting screen threshold.
# Read from the notebook's own analysis_params.csv when available, so the two
# cannot drift apart; otherwise fall back to the value 02b last reported.
params <- read_result("analysis_params.csv")
N_EFF_METABOLITES <- if (have(params)) params$n_eff_metabolites[1] else 1217   # Li & Ji.
p_thresh_mwas     <- if (have(params)) params$p_thresh_mwas[1]     else 0.05 / N_EFF_METABOLITES
# analysis_params.csv written before the switch to Li & Ji has no estimator column; those
# files used the participation ratio.
M_EFF_ESTIMATOR   <- if (have(params) && "m_eff_estimator" %in% names(params)) params$m_eff_estimator[1] else
                     if (have(params)) "participation ratio" else "Li & Ji (2005)"
threshold_source  <- if (have(params)) "read from analysis_params.csv" else "fallback default"

# Readable labels for outcomes and exposures.
outcome_labels  <- c(hdl_log = "log(HDL-C)", tg_log = "log(TG)", bmi = "BMI",
                     fg = "Fasting glucose", hb = "HbA1c")
exposure_labels <- c(mvpa_wins = "MVPA", bmi_covariate = "BMI", bmi = "BMI", sex = "Sex", smoking = "Smoking")

relabel <- function(x, lookup) coalesce(unname(lookup[x]), x)

# Display names for the QC tables. The raw name is the auditable key, so it stays
# in the exported CSVs; only what a reader sees is relabelled.
qc_var_labels <- c(
  age = "Age (years)", bmi = "BMI (kg/m2), as measured", hdl = "HDL-C (mg/dL)",
  hdl_log = "log(HDL-C)", tg = "Triglycerides (mg/dL)", tg_log = "log(TG)",
  bmi_covariate = "BMI (kg/m2), winsorized and imputed (exposure/covariate)",
  fg = "Fasting glucose (mg/dL)", hb = "HbA1c (%)",
  mvpa = "MVPA (MET-hours/week)", mvpa_wins = "MVPA, winsorized",
  mvpa_log = "log(MVPA + 1)", mvpa_trunc = "MVPA, truncated",
  pa = "Total PA", mod_pa = "Moderate PA", vig_pa = "Vigorous PA",
  ahei_score = "AHEI-2010 score", dash_score = "DASH score",
  ses_score = "Neighbourhood SES score", drinks_per_week = "Alcohol (drinks/week)",
  sex = "Sex", race = "Race/ethnicity (self-identified)", smoking = "Smoking",
  site = "Study site", income = "Income", season = "Season", month = "Month",
  insulin = "Insulin")

qc_label <- function(x) relabel(x, qc_var_labels)

# Harmonised in 01c but not analysed from 02a on, so kept out of every table and figure
# here even though 01c's QC exports still carry them.
not_analysed <- c("fg", "hb")

# Phenotype ICCs not shown: the other PA scales (MVPA is modelled as mvpa_wins), and
# traits no lane uses.
icc_not_shown <- c(not_analysed, "mvpa", "mvpa_log", "mvpa_trunc", "pa", "mod_pa", "vig_pa",
                   "tg_log", "ses_score")

qc_quantity_labels <- c(feature_missingness = "Missingness per feature",
                        sample_missingness  = "Missingness per sample",
                        qc_pool_cv          = "QC-pool CV")

# Anchor = one (outcome, exposure, SNP) combination taken into the screen.
# NB: the argument is `dat`, not `df` -- the results tables carry a column named
# `df` (Satterthwaite degrees of freedom), which would shadow the argument inside
# mutate()'s data mask and silently break the gene lookup.
anchor_label <- function(dat) {
  dat <- with_lane(dat)
  if ("lane_id" %in% names(dat) && !is.null(lanes_all))
    return(dat %>%
             left_join(select(lanes_all, lane_id, lane_label = label), by = "lane_id") %>%
             mutate(anchor = coalesce(lane_label, paste(G, Y, E, sep = " / "))) %>%
             select(-lane_label))
  has_gene <- "gene" %in% names(dat)
  dat %>%
    mutate(anchor = paste0(if (has_gene) coalesce(gene, rsid) else rsid,
                           " (", rsid, ")\n",
                           relabel(E, exposure_labels), " -> ",
                           relabel(Y, outcome_labels)))
}
```

```{r anchor-scope}
# ---------------------------------------------------------------------------
# Pre-specified anchor set.
#
# 02b screens a superset of (SNP, outcome, exposure) combinations. The anchors
# carried into the manuscript are chosen on prior grounds -- a published
# interaction we are deliberately following up -- not on how they rank in MESA,
# and they are declared in anchors.csv so the choice lives in one auditable
# place rather than being implied by a filter somewhere downstream.
#
# To change the reported scope, edit anchors.csv: flip `include` to FALSE to
# park an anchor (its rows stay in results/, they just stop being reported).
# Dropping the CETP locus to a physical-activity-only paper is one edit there.
# ---------------------------------------------------------------------------

anchors_cfg    <- read_config("anchors.csv")
lanes_all      <- anchors_cfg$data
anchors_source <- anchors_cfg$source
if (!is.null(lanes_all) && !("lane_id" %in% names(lanes_all))) {
  warning("anchors.csv has no lane_id column - publish the current version (see README).")
  lanes_all <- NULL
}
if (!is.null(lanes_all)) lanes_all <- mutate(lanes_all, include = as.logical(include))
anchors <- if (!is.null(lanes_all)) filter(lanes_all, include) else NULL

# The primary lane (first included row) defines the Table S1 sample; see 02a.
primary_lane <- if (!is.null(anchors) && nrow(anchors)) anchors[1, ] else NULL

# Results written before lanes existed carry (G, Y, E) but no lane_id; attach it by exact
# match. Rows matching no declared lane keep lane_id = NA and so fall out of scope.
# The trait a variable measures, ignoring how it was derived (bmi and bmi_covariate are
# both BMI). Same rule as the notebooks' trait().
trait <- function(v) sub("_(log|wins|trunc|covariate)$", "", v)

# Matched on the SNP and the outcome/exposure TRAITS, so a result written when the CETP
# exposure was still named bmi attaches to the lane that now names it bmi_covariate.
with_lane <- function(dat) {
  if (is.null(dat) || nrow(dat) == 0 || is.null(lanes_all)) return(dat)
  if (!("lane_id" %in% names(dat)) && all(c("G", "Y", "E") %in% names(dat)))
    dat <- dat %>%
      mutate(.yt = trait(Y), .et = trait(E)) %>%
      left_join(transmute(lanes_all, lane_id, G, .yt = trait(Y), .et = trait(E)),
                by = c("G", ".yt", ".et")) %>%
      select(-.yt, -.et)
  dat
}

if (is.null(anchors)) {
  warning("No usable anchors.csv (missing, or it predates lane_id) - reporting every lane present in the results.")
  in_scope    <- function(dat, ...) dat
  scope_note  <- "no usable anchors.csv (missing, or published before lane_id existed); every screened lane is reported"
} else {
  in_scope <- function(dat) {
    dat <- with_lane(dat)
    if (is.null(dat) || nrow(dat) == 0 || !("lane_id" %in% names(dat))) return(dat)
    filter(dat, lane_id %in% anchors$lane_id)
  }
  scope_note <- paste(anchors$label, collapse = "; ")
}

# Marginal tables (Y~G, E~G) are keyed by only part of a lane, and a pair two lanes share
# is written once per lane. Keep a pair if any included lane uses it, and show it once.
keep_pairs <- function(dat, a, b) {
  if (is.null(dat) || nrow(dat) == 0 || is.null(anchors)) return(dat)
  if ("lane_id" %in% names(dat))
    return(dat %>% filter(lane_id %in% anchors$lane_id) %>%
             distinct(across(-lane_id), .keep_all = TRUE))
  key <- distinct(tibble(.k = trait(anchors[[a]]), .g = anchors[[b]]))   # By trait, as with_lane().
  dat %>% mutate(.k = trait(.data[[a]]), .g = .data[[b]]) %>%
    semi_join(key, by = c(".k", ".g")) %>% select(-.k, -.g)
}

gxm_all    <- in_scope(gxm_all)
gxe        <- in_scope(gxe)
gxc        <- in_scope(gxc)
signif_gxm <- in_scope(signif_gxm)      # hit selection is per-row, so filtering
                                        # and re-selecting give the same rows.
y_g        <- keep_pairs(y_g, "Y", "G")
e_g        <- keep_pairs(e_g, "E", "G")
m_e        <- if (have(signif_gxm) && have(m_e)) {
                if ("lane_id" %in% names(m_e) && "lane_id" %in% names(signif_gxm)) {
                  semi_join(m_e, distinct(select(signif_gxm, lane_id, M)), by = c("lane_id", "M"))
                } else {                        # M_E.csv written before lane_id: match by trait.
                  m_e %>% mutate(.et = trait(E)) %>%
                    semi_join(distinct(transmute(signif_gxm, M, .et = trait(E))), by = c("M", ".et")) %>%
                    select(-.et)
                }
              } else m_e
```


# Input inventory

```{r input-inventory}
inputs %>%
  mutate(Loaded = ifelse(Loaded, "yes", "MISSING")) %>%
  kable(caption = "Result files available to this report") %>%
  kable_styling(full_width = FALSE)
```

Variant table read from: **`r voi_source`**

Lanes read from: **`r anchors_source`**

Lanes reported: **`r scope_note`**
(declared in `anchors.csv`; lanes with `include = FALSE` are still screened and remain
in `results/`, but are not reported here).

```{r lanes-declared}
if (!is.null(lanes_all)) {
  lanes_all %>%
    mutate(Outcome = relabel(Y, outcome_labels), Exposure = relabel(E, exposure_labels),
           Reported = ifelse(include, "yes", "no")) %>%
    select(Lane = lane_id, Label = label, Gene = gene, rsid, SNP = G, Outcome, Exposure,
           Reported, any_of(c(`Extra covariates` = "extra_covars")), Provenance = provenance) %>%
    kable(caption = paste("Declared lanes. Every row is screened by 02b; `Reported`",
                          "controls whether this report shows it.")) %>%
    kable_styling(full_width = FALSE)
}
```

Screen-wide significance threshold: **p < `r format(signif(p_thresh_mwas, 3), scientific = TRUE)`**
(Bonferroni over `r signif(N_EFF_METABOLITES, 4)` effective metabolites, `r M_EFF_ESTIMATOR`;
`r threshold_source`). Note that this corrects across metabolites **within a lane**,
not across the `r if (!is.null(lanes_all)) nrow(lanes_all) else "declared"` lanes,
each of which screens the same metabolome. Whether to also correct across lanes -- or
across distinct SNPs, since lanes can share one -- is an open threshold decision.

# Upstream QC

Everything in this section is rendered from `results/qc/`, written by the
preprocessing notebooks. Nothing here is recomputed, and no individual-level data
is read -- the point is that the decisions made before any model was fitted are
visible in the same document as the models, rather than only in the notebook that
made them.

## Sample construction

```{r qc-sample-flow}
if (have(qc$sample_flow)) {
  flow <- qc$sample_flow
  per_lane <- "lane_id" %in% names(flow)
  if (per_lane && !is.null(anchors)) flow <- filter(flow, lane_id %in% anchors$lane_id)
  if (per_lane && !is.null(lanes_all))
    flow <- left_join(flow, select(lanes_all, lane_id, Lane = label), by = "lane_id")
  if ("variable" %in% names(flow))   # Name the lane's own outcome/exposure in the stage.
    flow <- mutate(flow, stage = ifelse(is.na(variable) | grepl("^chr", variable), stage,
                                        paste0(stage, ": ", qc_label(variable))))

  flow_tbl <- flow %>%
    filter(exam == "all") %>%
    arrange(across(any_of("Lane")), step) %>%
    select(any_of("Lane"), Step = step, Stage = stage, Rows = n_rows,
           Participants = n_individuals, `Rows lost` = rows_dropped,
           `Participants lost` = individuals_dropped, Reason = reason)

  write_csv(flow_tbl, "output/manuscript/qc_sample_flow.csv")

  flow_tbl %>%
    kable(caption = paste("Sample flow. Filters are cumulative and in the order the",
                          "analysis applies them; a row is one participant-exam",
                          "observation. Steps from the SNP onward are each lane's own,",
                          "so the last row of each lane is the n that lane's model sees.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/sample_flow-mesa.csv")
```

```{r qc-sample-flow-by-exam}
if (have(qc$sample_flow)) {
  flow %>%
    filter(exam != "all") %>%
    arrange(across(any_of("Lane")), step) %>%
    select(any_of("Lane"), Stage = stage, exam, n_rows) %>%
    pivot_wider(names_from = exam, names_prefix = "Exam ", values_from = n_rows) %>%
    kable(caption = paste("Sample flow by exam. Exam 4 carries no physical-activity",
                          "questionnaire, and metabolomics was run on a subset of exams,",
                          "so the losses are not spread evenly -- and differ by lane.")) %>%
    kable_styling(full_width = FALSE)
}
```

```{r qc-method-coverage}
if (have(qc$metabolomics_method_coverage)) {
  qc$metabolomics_method_coverage %>%
    filter(exam == "all") %>%
    select(`LC-MS method` = method, Features = n_features,
           Observations = n_rows, Participants = n_individuals) %>%
    kable(caption = paste("Coverage by LC-MS method. A sample can pass QC on one",
                          "method and fail on another, so the n behind a given",
                          "metabolite depends on which method it came from.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/metabolomics_method_coverage-mesa.csv")
```

## Imputation

```{r qc-imputation}
if (have(qc$imputation_ledger)) {
  imp_tbl <- qc$imputation_ledger %>%
    filter(exam == "all", n_missing_before > 0, !variable %in% not_analysed) %>%
    arrange(desc(n_missing_before)) %>%
    select(Variable = variable, Imputation = imputation, Rows = n_rows,
           Missing = n_missing_before, `% missing` = frac_missing_before,
           Imputed = n_imputed, `Still missing` = n_still_missing) %>%
    mutate(`% missing` = signif(100 * `% missing`, 3))

  write_csv(imp_tbl, "output/manuscript/qc_imputation_ledger.csv")

  imp_tbl %>%
    kable(caption = paste("Imputation ledger, pooled over exams. Counts come from the",
                          "`__imp` flags 01c sets before any filling, so they record what",
                          "was actually missing. Every imputed value is one the models",
                          "treat as observed.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/imputation_ledger-mesa.csv")
```

## Exposure, outcome and covariate distributions

```{r qc-mvpa-transforms}
if (have(qc$variable_distributions)) {
  qc$variable_distributions %>%
    filter(grepl("^mvpa", variable), exam == "all") %>%     # Pooled over exams.
    mutate(variable = qc_label(variable)) %>%
    select(Variable = variable, n, `% zero` = n_zero, Mean = mean,
           SD = sd, Min = min, P25 = p25, Median = median, P75 = p75, Max = max) %>%
    mutate(`% zero` = signif(100 * `% zero` / n, 3),
           across(c(Mean, SD, Min, P25, Median, P75, Max), ~ signif(.x, 3))) %>%
    kable(caption = paste("MVPA under each transformation 02a compared, pooled over exams. `mvpa_wins` is",
                          "the one the models use: winsorized at the 90th percentile,",
                          "which is what moves the maximum.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/variable_distributions-mesa.csv")
```

## Within-person reliability (ICC)

The simulation assumes ICC = 0.5 for both the exposure and the mediator. These are
the same quantity measured in MESA, which is what connects the two halves of the
paper: if the real ICCs sit well below the assumed value, the simulation's power
curves are optimistic for this dataset.

Read them as a **lower bound** on reliability. The exams are 5-10 years apart, so
the within-person term mixes measurement error with genuine change, and anything
that really moves over a decade -- MVPA above all -- is penalised for moving.

```{r qc-icc, fig.asp = 0.5}
if (have(qc$icc) || have(qc$metabolite_icc_summary)) {
  # One frame for both halves of the panel: phenotypes are single estimates,
  # metabolites are a distribution over features, and they share the y axis.
  icc_panel <- bind_rows(
    if (have(qc$icc))
      qc$icc %>% filter(!variable %in% icc_not_shown) %>% arrange(desc(icc)) %>%
        transmute(label = qc_label(variable), kind = "Phenotype", point = icc) else NULL,
    if (have(qc$metabolite_icc_summary))
      qc$metabolite_icc_summary %>%
        transmute(label = gsub("_", "-", method), kind = "Metabolites (per feature)",
                  p05, p25, median, p75, p95) else NULL) %>%
    mutate(label = fct_inorder(label),
           kind  = factor(kind, levels = c("Phenotype", "Metabolites (per feature)")))

  ggplot(icc_panel, aes(x = label)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "firebrick") +
    { if ("median" %in% names(icc_panel))
        geom_boxplot(aes(ymin = p05, lower = p25, middle = median, upper = p75,
                         ymax = p95),
                     stat = "identity", width = 0.55, fill = "grey85", na.rm = TRUE) } +
    { if ("point" %in% names(icc_panel))
        geom_point(aes(y = point), size = 2, na.rm = TRUE) } +
    facet_grid(cols = vars(kind), scales = "free_x", space = "free_x") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = NULL, y = "ICC across exams",
         caption = paste("Dashed line: the ICC = 0.5 the simulation assumes.",
                         "Boxes are the P25-P75 of the per-feature ICCs, whiskers P05-P95.")) +
    scale_x_discrete(labels = scales::label_wrap(11)) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
} else skip_note("qc/icc-mesa.csv", "qc/metabolite_icc_summary-mesa.csv")
```

```{r qc-icc-table}
if (have(qc$icc)) {
  icc_tbl <- qc$icc %>%
    filter(!variable %in% icc_not_shown) %>%
    arrange(desc(icc)) %>%
    mutate(variable = qc_label(variable)) %>%
    select(Variable = variable, Participants = n_individuals,
           Observations = n_observations, `Between-person var` = var_between,
           `Within-person var` = var_within, ICC = icc) %>%
    mutate(across(where(is.numeric) & !c(Participants, Observations), ~ signif(.x, 3)))

  write_csv(icc_tbl, "output/manuscript/qc_icc.csv")

  icc_tbl %>%
    kable(caption = paste("Within-person ICC(1) across exams, among participants with",
                          "at least two measured values. Carried-forward values are",
                          "excluded: they are copies of the previous exam's and would",
                          "drive the ICC towards 1.")) %>%
    kable_styling(full_width = FALSE)
}
```

```{r qc-icc-hits}
# The hits' own reliability is the part of this that bears on the results: a hit on
# a feature with near-zero ICC is a hit on something the assay barely reproduces.
if (have(qc$metabolite_icc) && have(signif_gxm)) {
  hit_icc <- signif_gxm %>%
    select(M, met_label) %>%
    distinct() %>%
    inner_join(select(qc$metabolite_icc, M = feature, icc, n_individuals), by = "M")

  if (nrow(hit_icc)) {
    hit_icc %>%
      arrange(desc(icc)) %>%
      select(Metabolite = met_label, Participants = n_individuals, ICC = icc) %>%
      mutate(ICC = signif(ICC, 3)) %>%
      kable(caption = "Within-person ICC of the significant metabolites") %>%
      kable_styling(full_width = FALSE)
  } else cat("No overlap between the hits and the metabolites with repeated measures.\n")
}
```

## Covariate justification

```{r qc-covariate-ladder, fig.asp = 0.6}
if (have(qc$covariate_ladder)) {
  qc$covariate_ladder %>%
    filter(outcome %in% (if (!is.null(anchors)) anchors$Y else "hdl_log"),
           !grepl("^chr", variable),
           # +bmi repeats the rung before it when BMI is already the outcome or the variable.
           !(covar_set == "+bmi" & (trait(outcome) == "bmi" | trait(variable) == "bmi"))) %>%
    mutate(covar_set = fct_inorder(covar_set), variable = qc_label(variable),
           outcome = qc_label(outcome)) %>%
    ggplot(aes(x = covar_set, y = estimate, color = covar_set)) +
    geom_hline(yintercept = 0, color = "gray70") +
    geom_pointrange(aes(ymin = l95, ymax = u95), size = 0.3) +
    facet_wrap(vars(outcome, variable), scales = "free_y",
               labeller = label_wrap_gen(width = 28, multi_line = TRUE)) +
    scale_x_discrete(labels = NULL, breaks = NULL) +
    labs(x = NULL, y = "Standardised estimate",
         color = "Covariate set",
         caption = "Rungs are cumulative; the last is the set 02b fits.")
} else skip_note("qc/covariate_ladder-mesa.csv")
```

```{r qc-variance-explained, fig.asp = 0.45}
if (have(qc$covariate_variance_explained)) {
  ve <- qc$covariate_variance_explained %>%
    filter(!(covar_set == "+bmi" & trait(outcome) == "bmi"),     # A no-op rung for BMI itself.
           outcome %in% (if (!is.null(anchors)) c(anchors$Y, anchors$E) else outcome)) %>%
    # Levels from the whole file: an outcome that skips a rung would otherwise reorder them.
    mutate(covar_set = factor(covar_set, levels = unique(qc$covariate_variance_explained$covar_set)),
           outcome = qc_label(outcome))

  write_csv(ve, "output/manuscript/qc_covariate_variance_explained.csv")

  ggplot(ve, aes(x = covar_set, y = adj_r2, color = outcome, group = outcome)) +
    geom_line() +
    geom_point(size = 1.8) +
    scale_color_discrete(labels = scales::label_wrap(30)) +
    labs(x = "Covariate set (cumulative; -race drops race but keeps gPCs)",
         y = expression("Adjusted " * R^2), color = NULL,
         caption = "All rungs fitted on the same complete-case rows, so they are comparable.") +
    theme(axis.text.x = element_text(angle = 35, hjust = 0.9), legend.position = "right")
} else skip_note("qc/covariate_variance_explained-mesa.csv")
```

Each cell below is one `mPC ~ covariate` linear model. Only exam-1 samples with
data on all four LC-MS methods contribute to the metabolite PCs, so this is a
statement about that subset rather than about the whole analysis sample.

```{r qc-mpc-heatmap, fig.asp = 0.6}
if (have(qc$mpc_covariate_pvalues)) {
  qc$mpc_covariate_pvalues %>%
    filter(as.integer(sub("mPC", "", mPC)) <= 15) %>%            # 02a's own heatmap shows 15.
    mutate(mPC = factor(mPC, levels = paste0("mPC", 1:100)),
           nlp = pmin(-log10(p), 10)) %>%
    ggplot(aes(x = mPC, y = fct_rev(fct_inorder(covariate)), fill = nlp)) +
    geom_tile() +
    geom_text(data = ~ filter(.x, p < 0.05), label = "*", color = "white", size = 2.5,
              vjust = 0.75) +
    scale_fill_gradient(name = expression(-log[10](italic(p))),
                        low = "black", high = "steelblue", limits = c(0, 10)) +
    labs(x = NULL, y = NULL, caption = "* marks p < 0.05. Shading capped at 10.") +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9))
} else skip_note("qc/mpc_covariate_pvalues-mesa.csv")
```

## Metabolite QC

```{r qc-met-steps}
if (have(qc$metabolite_qc_steps)) {
  steps_tbl <- qc$metabolite_qc_steps %>%
    # Step index and label are separate columns: a label starting "1." is read as
    # an ordered-list marker inside a table cell and swallowed by pandoc.
    { if ("step_label" %in% names(.)) . else mutate(., step_label = as.character(step)) } %>%
    select(Method = method_label, Step = step, Stage = step_label,
           Features = n_features, Samples = n_samples,
           `Features dropped` = features_dropped, `Samples dropped` = samples_dropped)

  write_csv(steps_tbl, "output/manuscript/qc_metabolite_steps.csv")

  steps_tbl %>%
    kable(caption = paste("Features and samples surviving each 01a QC step. The last",
                          "row is the merged matrix: a sample profiled by only some",
                          "methods still contributes, which is why it has more samples",
                          "than any single method.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/metabolite_qc_steps-mesa.csv")
```

```{r qc-met-distributions, fig.asp = 0.4}
if (have(qc$metabolite_distributions)) {
  qc$metabolite_distributions %>%
    # The three quantities use different breaks, so the bins are ordered by their
    # lower edge rather than by order of appearance in the file.
    mutate(quantity = factor(recode(quantity, !!!qc_quantity_labels),
                             levels = unname(qc_quantity_labels)),
           bin = fct_reorder(bin, bin_lower)) %>%
    ggplot(aes(x = bin, y = n, fill = method_label)) +
    geom_col(position = "dodge") +
    facet_wrap(vars(quantity), scales = "free") +
    labs(x = NULL, y = "Count", fill = "Method") +
    theme(axis.text.x = element_text(angle = 40, hjust = 0.9))
} else skip_note("qc/metabolite_distributions-mesa.csv")
```

```{r qc-met-quantiles}
if (have(qc$metabolite_qc_quantiles)) {
  qc$metabolite_qc_quantiles %>%
    mutate(quantity = recode(quantity, !!!qc_quantity_labels),
           across(c(mean, p05, p25, median, p75, p95), ~ signif(.x, 3))) %>%
    select(Method = method_label, Quantity = quantity, n, Mean = mean,
           P05 = p05, P25 = p25, Median = median, P75 = p75, P95 = p95,
           any_of(c(`QC-pool samples` = "n_qc_pool_samples"))) %>%
    kable(caption = paste("Missingness and pooled-QC CV. The CV is taken on the raw",
                          "intensity scale before the log2, over the pooled QC samples",
                          "01a drops, so it measures analytical reproducibility rather",
                          "than biological spread.")) %>%
    kable_styling(full_width = FALSE)
}
```

### The screen threshold

```{r qc-meff}
if (have(qc$m_eff_estimators)) {
  meff_tbl <- qc$m_eff_estimators %>%
    mutate(Used = ifelse(in_use, "yes", ""),
           m_eff = signif(m_eff, 4), p_threshold = signif(p_threshold, 3)) %>%
    select(Estimator = estimator, `M_eff` = m_eff, `p threshold` = p_threshold, Used)

  write_csv(meff_tbl, "output/manuscript/qc_m_eff.csv")

  meff_tbl %>%
    kable(caption = paste("Effective number of independent metabolites under several",
                          "estimators, over", qc$m_eff_estimators$n_features[1],
                          "features in", qc$m_eff_estimators$n_samples[1],
                          "samples with complete data on all four methods. The screen",
                          "uses Li & Ji; the simulated max statistic is the threshold that",
                          "holds the family-wise error at 0.05 directly, assuming the GxM",
                          "statistics are correlated as the features are.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/m_eff_estimators-mesa.csv")
```

```{r qc-eigenvalues, fig.asp = 0.42}
if (have(qc$metabolite_eigenvalues)) {
  spec <- qc$metabolite_eigenvalues

  p_spec <- spec %>%
    filter(component <= 200) %>%
    ggplot(aes(x = component, y = eigenvalue)) +
    geom_line() +
    scale_y_log10() +
    labs(x = "Component", y = "Eigenvalue (log scale)")

  p_cum <- spec %>%
    ggplot(aes(x = component, y = cum_var_explained)) +
    geom_line() +
    geom_hline(yintercept = 0.995, linetype = "dashed", color = "gray50") +   # simpleM.
    { if (have(qc$m_eff_estimators))
        geom_vline(data = filter(qc$m_eff_estimators, in_use),
                   aes(xintercept = m_eff), color = "firebrick") } +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Component", y = "Cumulative variance explained",
         caption = "Red line: the M_eff the screen threshold uses.")

  p_spec + p_cum
} else skip_note("qc/metabolite_eigenvalues-mesa.csv")
```

## Genotype QC

```{r qc-genotype}
if (have(qc$genotype_qc)) {
  geno_tbl <- qc$genotype_qc %>%
    { if (!is.null(lanes_all)) filter(., cpaid %in% lanes_all$G) %>%
        left_join(distinct(select(lanes_all, cpaid = G, .gene = gene, .rsid = rsid)), by = "cpaid") %>%
        mutate(gene = coalesce(na_if(gene, ""), .gene), rsid = coalesce(na_if(rsid, ""), .rsid)) %>%
        select(-.gene, -.rsid) else . } %>%
    # 01b names each dosage column chr_pos_ref_alt and polarizes it to count that alt, so
    # the counted allele is the cpaid's last field. An older 01b export joined its
    # annotation on a cpaid that no longer matched, and recorded the allele as undefined.
    mutate(.alt = sub(".*_", "", cpaid),
           counted_allele = ifelse(grepl("^UNDEFINED", counted_allele) & nzchar(.alt),
                                   paste0("alt = ", .alt), counted_allele)) %>%
    select(-.alt) %>%
    mutate(across(c(alt_freq, maf, call_rate), ~ signif(.x, 3))) %>%
    select(any_of(c("gene", "rsid", "cpaid")), `Counted allele` = counted_allele,
           `Freq (counted)` = alt_freq, MAF = maf, Called = n_called,
           `Call rate` = call_rate)

  write_csv(geno_tbl, "output/manuscript/qc_genotype.csv")

  geno_tbl %>%
    kable(caption = paste("Call rate and dosage polarity for every declared lane's SNP.",
                          "01b re-polarizes a flipped variant so every dosage counts the",
                          "allele named in its `cpaid`; a variant whose counted allele is",
                          "undefined has no named alt in variants_of_interest.csv, and",
                          "the sign of its effect is relative to an unlabelled allele.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/genotype_qc-mesa.csv")
```

```{r qc-genotype-by-race}
if (have(qc$genotype_by_race)) {
  qc$genotype_by_race %>%
    { if (!is.null(lanes_all)) filter(., variant %in% lanes_all$G, !is.na(race)) else . } %>%
    mutate(across(c(freq_counted, maf, call_rate), ~ signif(.x, 3))) %>%
    select(Variant = variant, Race = race, Participants = n_participants,
           Called = n_called, `Freq (counted)` = freq_counted, MAF = maf,
           `Call rate` = call_rate) %>%
    kable(caption = paste("Allele frequency by self-identified race. MESA is",
                          "multi-ethnic, so a frequency that swings across groups is",
                          "the reason the models adjust for genetic PCs and fit",
                          "gPC x E interaction terms.")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/genotype_by_race-mesa.csv")
```

## Table S1. Participant characteristics

```{r qc-table-s1}
if (have(qc$table_s1_characteristics)) {
  s1 <- qc$table_s1_characteristics %>%
    mutate(cell = case_when(
      variable == "N"   ~ as.character(n),
      !is.na(level)     ~ paste0(n, " (", sprintf("%.1f", pct), "%)"),
      TRUE              ~ paste0(signif(mean, 3), " ± ", signif(sd, 2))),
      row = ifelse(is.na(level), qc_label(variable),
                   paste0(qc_label(variable), ": ", level))) %>%
    select(row, group, cell) %>%
    pivot_wider(names_from = group, values_from = cell) %>%
    # A cell is empty only when that level does not occur in that stratum.
    mutate(across(-row, ~ replace_na(.x, "-")))

  write_csv(s1, "output/manuscript/table_s1_characteristics.csv")

  s1 %>%
    rename(` ` = row) %>%
    kable(caption = paste0("Table S1. Characteristics of the primary analysis sample ",
                           "(exam-1 rows with metabolomics, genotype at the primary lane's SNP",
                           if (!is.null(primary_lane)) paste0(" (", primary_lane$gene, " ", primary_lane$rsid, ")") else "",
                           " and its outcome",
                           if (!is.null(primary_lane)) paste0(", ", qc_label(primary_lane$Y), ",") else "",
                           " measured). Continuous variables are mean +/- SD, ",
                           "categorical are n (%).")) %>%
    kable_styling(full_width = FALSE)
} else skip_note("qc/table_s1_characteristics-mesa.csv")
```

```{r qc-n-by-exam}
if (have(qc$analysis_n_by_exam)) {
  qc$analysis_n_by_exam %>%
    select(Exam = exam, Observations = n_rows, Participants = n_individuals) %>%
    kable(caption = paste("Observations contributed by each exam under the same",
                          "filters. The longitudinal models pool these, which is why",
                          "the model n exceeds the number of participants.")) %>%
    kable_styling(full_width = FALSE)
}
```

# Stage 0: how do the anchoring interactions look in MESA?

The anchors are chosen on prior grounds, so these tables are reported for
transparency rather than as a gate: they show how strongly each published
interaction is expressed in MESA under the current model specification, before
any metabolite is examined. A weak anchor is informative in its own right --
the simulation predicts that the GxM test is least specific exactly there.

```{r anchor-gxe}
if (have(gxe)) {
  gxe_tbl <- gxe %>%
    filter(!is.na(est)) %>%
    { if ("lane_id" %in% names(.) && !is.null(lanes_all))
        left_join(., select(lanes_all, lane_id, Lane = label), by = "lane_id") else . } %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene),
                  by = "G") else . } %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    arrange(p) %>%
    select(any_of(c("Lane", "Outcome", "Exposure", "rsid", "gene")),
           beta_GxE = est, se, p, n) %>%
    mutate(across(c(beta_GxE, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(gxe_tbl, "output/manuscript/anchor_gxe.csv")

  gxe_tbl %>%
    kable(caption = "Anchoring GxE interactions in MESA") %>%
    kable_styling(full_width = FALSE)
} else cat("GxE.csv not available - skipping.\n")
```

```{r anchor-marginal}
if (have(y_g) || have(e_g)) {
  marg <- bind_rows(
    if (have(y_g)) mutate(filter(y_g, !is.na(est)), Model = "Y ~ G",
                          Trait = relabel(Y, outcome_labels)) else NULL,
    if (have(e_g)) mutate(filter(e_g, !is.na(est)), Model = "E ~ G",
                          Trait = relabel(E, exposure_labels)) else NULL
  ) %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    arrange(p) %>%
    select(Model, Trait, any_of(c("rsid", "gene")), beta = est, se, p, n) %>%
    mutate(across(c(beta, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(marg, "output/manuscript/anchor_marginal.csv")

  marg %>%
    head(30) %>%
    kable(caption = paste("Marginal SNP associations (top 30). E ~ G is the",
                          "diagnostic for an upstream (scenario B) architecture:",
                          "a strong G -> E effect would suggest the SNP acts on",
                          "the exposure rather than downstream of it.")) %>%
    kable_styling(full_width = FALSE)
} else cat("Y_G.csv / E_G.csv not available - skipping.\n")
```

# Metabolome-wide interaction screen

```{r mwis-counts}
if (have(gxm_all)) {
  gxm_annot <- gxm_all %>%
    filter(!is.na(est)) %>%
    { if (have(met_info))
        left_join(., select(met_info, M = unique_met_id, Metabolite, met_label,
                            Method, MZ, RT, HMDB_ID),
                  by = "M") else . } %>%
    { if (!("met_label" %in% names(.)) && have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    mutate(hit = p < p_thresh_mwas)

  screen_summary <- gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    anchor_label() %>%
    group_by(Lane = anchor, across(any_of(c("rsid", "gene"))), Y, E) %>%
    summarise(`Metabolites tested` = n(),
              `Hits` = sum(hit),
              `Min p` = signif(min(p, na.rm = TRUE), 3),
              .groups = "drop") %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    select(Lane, any_of(c("rsid", "gene")), Outcome, Exposure,
           `Metabolites tested`, Hits, `Min p`) %>%
    arrange(desc(Hits))

  write_csv(screen_summary, "output/manuscript/screen_summary.csv")

  screen_summary %>%
    kable(caption = "Metabolome-wide screen: hits per lane") %>%
    kable_styling(full_width = FALSE)
} else cat("GxM_results.csv not available - skipping.\n")
```

```{r mwis-volcano, fig.asp = 0.5}
if (have(gxm_all)) {
  volcano_df <- gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    anchor_label()

  volcano_plt <- volcano_df %>%
    ggplot(aes(x = est, y = -log10(p))) +
    geom_point(aes(color = hit), size = 0.7, alpha = 0.6) +
    geom_hline(yintercept = -log10(p_thresh_mwas), linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, color = "gray80") +
    { if ("met_label" %in% names(volcano_df))
        geom_text_repel(data = filter(volcano_df, hit), aes(label = met_label),
                        size = 2.2, min.segment.length = 0, box.padding = 0.6,
                        point.padding = 0.2, segment.size = 0.3, segment.color = "gray50",
                        max.overlaps = Inf, seed = 1, direction = "y", hjust = 0) } +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.28))) +
    scale_color_manual(values = c(`FALSE` = "gray70", `TRUE` = "firebrick"),
                       guide = "none") +
    facet_wrap(vars(anchor), scales = "free_x") +
    labs(x = expression("G x M interaction estimate (" * beta[GxM] * ")"),
         y = expression(-log[10](italic(p))))

  volcano_plt
} else cat("GxM_results.csv not available - skipping.\n")
```

```{r mwis-by-method, fig.asp = 0.45}
if (have(gxm_all) && "Method" %in% names(gxm_annot)) {
  gxm_annot %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    anchor_label() %>%
    arrange(Method) %>%
    group_by(anchor) %>%
    mutate(idx = row_number()) %>%
    ungroup() %>%
    ggplot(aes(x = idx, y = -log10(p), color = Method)) +
    geom_point(size = 0.6, alpha = 0.7) +
    geom_hline(yintercept = -log10(p_thresh_mwas), linetype = "dashed", color = "gray40") +
    facet_wrap(vars(anchor)) +
    labs(x = "Metabolite (ordered by LC-MS method)",
         y = expression(-log[10](italic(p)))) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
} else cat("Metabolite metadata (Method) not available - skipping per-method plot.\n")
```

```{r signif-hits-table}
if (have(signif_gxm)) {
  hits_tbl <- signif_gxm %>%
    { if (!("label" %in% names(.)) && "lane_id" %in% names(.) && !is.null(lanes_all))
        left_join(., select(lanes_all, lane_id, label), by = "lane_id") else . } %>%   # Hits written before lanes.
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels)) %>%
    arrange(p) %>%
    select(any_of(c(Lane = "label", "rsid", "gene")), Outcome, Exposure,
           Metabolite = met_label, any_of(c("HMDB_ID", "Method", "MZ", "RT")),
           beta_GxM, se, p) %>%
    mutate(across(any_of(c("MZ")), ~ signif(.x, 6)),
           across(any_of(c("RT")), ~ signif(.x, 3)),
           across(c(beta_GxM, se, p), ~ signif(.x, 3)))

  write_csv(hits_tbl, "output/manuscript/table1_gxm_hits.csv")

  hits_tbl %>%
    kable(caption = "Table 1. Significant gene x metabolite interactions") %>%
    kable_styling(full_width = FALSE)
} else cat("mesa_signif_GxMs.csv not available - skipping.\n")
```

# Directness: how much interaction signal does GxM recover that GxE does not?

The conceptual model says the observed GxE interaction is a *diluted* version of
the GxM interaction: with E and M standardised and coupled by
`M = alpha*E + eps`, we expect `beta_GxE ~= alpha * beta_GxM`. Dilution has two
sources -- imperfect measurement of E, and the structural fact that M has
determinants other than E -- and both push `alpha` below 1 regardless of how well
E is measured. The tables below quantify that gap directly in the MESA data.

```{r gxm-vs-gxe}
if (have(signif_gxm) && have(gxe)) {
  gxe_key <- gxe %>%
    filter(!is.na(est)) %>%
    select(Y, E, G, beta_GxE = est, se_GxE = se, p_GxE = p)

  directness <- signif_gxm %>%
    inner_join(gxe_key, by = c("Y", "E", "G")) %>%
    mutate(Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels),
           ratio_beta = beta_GxM / beta_GxE,
           gain_nlp = -log10(p) - (-log10(p_GxE))) %>%
    arrange(p) %>%
    select(any_of(c("rsid", "gene")), Outcome, Exposure, Metabolite = met_label,
           beta_GxM, p_GxM = p, beta_GxE, p_GxE,
           `beta ratio (GxM/GxE)` = ratio_beta,
           `-log10p gain` = gain_nlp) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(directness, "output/manuscript/directness_gxm_vs_gxe.csv")

  directness %>%
    kable(caption = paste("GxM versus the anchoring GxE. A beta ratio > 1 and a",
                          "positive -log10p gain indicate the metabolite carries",
                          "the interaction more directly than the exposure does.")) %>%
    kable_styling(full_width = FALSE)
} else cat("Need both mesa_signif_GxMs.csv and GxE.csv - skipping.\n")
```

```{r dilution-check, fig.asp = 0.5, results = 'asis'}
# Quantitative test of the conceptual model: predicted beta_GxE = alpha * beta_GxM,
# where alpha is the standardised M ~ E coefficient. Deviations are informative --
# they bound how much of the GxE signal the exposure can carry at all.
if (have(signif_gxm) && have(gxe) && have(m_e)) {
  alpha_df <- m_e %>%
    filter(!is.na(est)) %>%
    transmute(M, .et = trait(E), alpha = est, alpha_se = se, alpha_p = p) %>%
    distinct(M, .et, .keep_all = TRUE)  # One alpha per (metabolite, exposure trait); guards the join below.

  dilution <- signif_gxm %>%
    inner_join(select(filter(gxe, !is.na(est)), lane_id, beta_GxE = est),
               by = "lane_id") %>%
    mutate(.et = trait(E)) %>%
    inner_join(alpha_df, by = c("M", ".et")) %>%
    mutate(predicted_GxE = alpha * beta_GxM,
           Outcome = relabel(Y, outcome_labels),
           Exposure = relabel(E, exposure_labels))

  if (nrow(dilution) > 0) {
    dil_tbl <- dilution %>%
      arrange(p) %>%
      select(any_of(c("rsid", "gene")), Outcome, Exposure, Metabolite = met_label,
             `alpha (M~E)` = alpha, beta_GxM, `predicted beta_GxE` = predicted_GxE,
             `observed beta_GxE` = beta_GxE) %>%
      mutate(across(where(is.numeric), ~ signif(.x, 3)))

    write_csv(dil_tbl, "output/manuscript/dilution_check.csv")

    cat("\n\n")
    print(kable_styling(kable(
      dil_tbl,
      caption = paste("Dilution check: is the observed GxE consistent with",
                      "alpha * beta_GxM? alpha is the standardised metabolite-exposure",
                      "coupling, so a small alpha implies the exposure can only ever",
                      "carry a fraction of the interaction."))))
    cat("\n\n")

    dilution %>%
      ggplot(aes(x = predicted_GxE, y = beta_GxE)) +
      geom_abline(slope = 1, intercept = 0, color = "gray60", linetype = "dashed") +
      geom_hline(yintercept = 0, color = "gray85") +
      geom_vline(xintercept = 0, color = "gray85") +
      geom_point(aes(color = Exposure), size = 2) +
      geom_text_repel(aes(label = met_label), size = 2.5, max.overlaps = 15) +
      labs(x = expression("Predicted " * beta[GxE] * " ( = " * alpha %*% beta[GxM] * " )"),
           y = expression("Observed " * beta[GxE]))
  } else cat("No overlapping (M, E) rows between the hits and M_E.csv - skipping.\n")
} else cat("Need mesa_signif_GxMs.csv, GxE.csv and M_E.csv - skipping.\n")
```

# Are the hits plausible mediators?

A metabolite that mediates the exposure's contribution should itself be
associated with the exposure. Metabolites with a strong GxM but no M~E
association are candidates for the reverse-causation or confounding
architectures rather than true mediation.

```{r m-e-table}
if (have(m_e)) {
  me_tbl <- m_e %>%
    filter(!is.na(est)) %>%
    { if (have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    { if (have(met_info) && !("met_label" %in% names(.)))
        left_join(., select(met_info, M = unique_met_id, met_label), by = "M") else . } %>%
    mutate(Exposure = relabel(E, exposure_labels)) %>%
    # any_of() on both names would keep both, as Metabolite1/Metabolite2; M is the fallback.
    mutate(Metabolite = if ("met_label" %in% names(.)) coalesce(met_label, M) else M) %>%
    arrange(p) %>%
    select(Metabolite, Exposure,
           beta_E = est, se, p, n) %>%
    mutate(across(c(beta_E, se), ~ signif(.x, 3)), p = signif(p, 3))

  write_csv(me_tbl, "output/manuscript/metabolite_exposure.csv")

  me_tbl %>%
    kable(caption = "Metabolite-exposure associations for the significant hits") %>%
    kable_styling(full_width = FALSE)
} else cat("M_E.csv not available - skipping.\n")
```

# Redundancy among the hits

Significant metabolites are often strongly correlated with one another, so a raw
count of hits overstates the number of independent signals. This section
collapses them where the QC'd metabolite matrix is available locally.

```{r hit-correlations, fig.asp = 0.8}
met_matrix_path <- "data/derived/metabolomics/QCd/merged_QCd-mesa.csv"

if (have(signif_gxm) && file.exists(met_matrix_path)) {
  hit_ids <- unique(signif_gxm$M)
  met_mat <- read_csv(met_matrix_path, show_col_types = FALSE, progress = FALSE) %>%
    select(any_of(hit_ids))

  if (ncol(met_mat) > 1) {
    cmat <- cor(met_mat, use = "pairwise.complete.obs")
    lab <- signif_gxm$met_label[match(colnames(cmat), signif_gxm$M)]
    dimnames(cmat) <- list(lab, lab)

    clust <- hclust(as.dist(1 - abs(cmat)))
    n_indep <- length(unique(cutree(clust, h = 1 - 0.8)))  # Merge at |r| > 0.8

    cat("Independent signal clusters at |r| < 0.8:", n_indep,
        "from", ncol(cmat), "significant metabolites\n\n")

    as_tibble(cmat, rownames = "met1") %>%
      pivot_longer(-met1, names_to = "met2", values_to = "r") %>%
      mutate(across(c(met1, met2), ~ factor(.x, levels = clust$labels[clust$order]))) %>%
      ggplot(aes(x = met1, y = met2, fill = r)) +
      geom_tile() +
      scale_fill_gradient2(name = "Pearson r", limits = c(-1, 1)) +
      labs(x = "", y = "") +
      theme(axis.text.x = element_text(angle = 40, hjust = 0.9))
  } else cat("Fewer than two significant metabolites - nothing to cluster.\n")
} else {
  cat("Metabolite matrix not available locally (", met_matrix_path, ").\n",
      "Copy it down, or run this section on Terra, to collapse correlated hits.\n", sep = "")
}
```

# Sensitivity: G x covariate adjustment

```{r gxcovars}
if (have(gxc)) {
  gxc_tbl <- gxc %>%
    filter(!is.na(est)) %>%
    { if (have(signif_gxm))
        left_join(., distinct(select(signif_gxm, M, met_label)), by = "M") else . } %>%
    { if (!is.null(variants_of_interest))
        left_join(., select(variants_of_interest, G = cpaid, rsid, gene), by = "G") else . } %>%
    mutate(Outcome = relabel(Y, outcome_labels)) %>%
    mutate(Metabolite = if ("met_label" %in% names(.)) coalesce(met_label, M) else M) %>%
    arrange(p) %>%
    select(any_of(c("rsid", "gene")), Outcome, Metabolite,
           beta_GxM_adj = est, se, p) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(gxc_tbl, "output/manuscript/sensitivity_gxcovars.csv")

  gxc_tbl %>%
    kable(caption = paste("GxM estimates with additional G x covariate adjustment.",
                          "Estimates that attenuate substantially relative to Table 1",
                          "suggest the interaction is confounded rather than causal.")) %>%
    kable_styling(full_width = FALSE)
} else cat("GxMpGxEpGxC.csv not available - skipping.\n")
```

```{r attenuation-comparison}
if (have(gxc) && have(signif_gxm)) {
  atten <- signif_gxm %>%
    select(lane_id, M, met_label, beta_primary = beta_GxM, p_primary = p) %>%
    inner_join(select(filter(gxc, !is.na(est)), lane_id, M,     # By lane, not (Y, E, G), so an
                      beta_adj = est, p_adj = p),               # older file's E = bmi still joins.
               by = c("lane_id", "M")) %>%
    mutate(`% attenuation` = 100 * (1 - beta_adj / beta_primary)) %>%
    arrange(p_primary) %>%
    select(Metabolite = met_label, beta_primary, beta_adj,
           `% attenuation`, p_primary, p_adj) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3)))

  write_csv(atten, "output/manuscript/sensitivity_attenuation.csv")

  atten %>%
    kable(caption = "Primary versus G x covariate-adjusted GxM estimates") %>%
    kable_styling(full_width = FALSE)
}
```

# Simulation study

Operating characteristics of the screening pipeline, computed by
`04_simulation.ipynb` and read here from `results/simulation/`. The notebook
writes rejection rates and per-replicate p-values; every figure below is a
rendering of one of those tables, so a restyle or a re-facet costs a knit rather
than a re-run of the simulation.

```{r load-sim}
sim_dir <- file.path(results_dir, "simulation")

read_sim <- function(f) {
  path <- file.path(sim_dir, f)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE, progress = FALSE)
}

sim_params  <- read_sim("sim_params.csv")
sim_power   <- read_sim("power_summary.csv")
sim_power_z <- read_sim("power_z_replicates.csv")
sim_z_agree <- read_sim("power_z_summary.csv")
sim_bias    <- read_sim("power_bias_summary.csv")
sim_dil_rep <- read_sim("dilution_replicates.csv")
sim_dil     <- read_sim("dilution_summary.csv")
sim_medmod  <- read_sim("medmod_summary.csv")
sim_icc     <- read_sim("icc_grid_summary.csv")
sim_tie     <- read_sim("tie_summary.csv")
sim_tie_b   <- read_sim("tie_scenarioB_summary.csv")
sim_tie_c   <- read_sim("tie_scenarioC_summary.csv")
sim_nsweep  <- read_sim("tie_scenarioB_nsweep_summary.csv")
sim_gamma   <- read_sim("tie_gamma_summary.csv")
dag_nodes   <- read_sim("dag_nodes.csv")
dag_edges   <- read_sim("dag_edges.csv")

sim_param <- function(name, default = NA) {
  if (!have(sim_params)) return(default)
  v <- sim_params$value[sim_params$param == name]
  if (length(v) == 0) default else v[1]
}
SIM_ALPHA <- as.numeric(sim_param("ALPHA", 0.05))

# The exported tables carry raw keys, so every display label lives here. That
# keeps one styling layer for the whole report rather than one in the notebook
# and another here, which is what let the two drift apart previously.
test_labels <- c(p_gxe        = "GxE",         p_gxm     = "GxM",
                 p_med        = "Mediation",
                 p_unadj      = "Unadjusted",  unadj     = "Unadjusted",
                 p_adj_GxE    = "G:E-adjusted", adjGxE   = "G:E-adjusted",
                 p_adj_U_main = "U-adjusted",   adjU     = "U-adjusted",
                 p_adj_GxU    = "G:U-adjusted", adjGxU   = "G:U-adjusted")
variant_labels <- c(full         = "Baseline",
                    full_adj_GxE = "+ G:E adjustment",
                    no_marg      = "No marginal E-M correlation")
icc_labels  <- c(perfect = "Perfect measurement",
                 default = "Realistic measurement error")
medmod_labels <- c(perfect = "Perfect (ICC = 1)", realistic = "Realistic (ICC = 0.5)")
scenario_labels <- c(A = "Molecular mediator (primary)", B = "Upstream GxE",
                     C = "Reverse causation",            D = "E-M Confounding")

sim_cols   <- c(GxE = "#d95f02", GxM = "#1b9e77", Mediation = "#7570b3")
sim_shapes <- c(GxE = 17, GxM = 16, Mediation = 15)
D_MODELS   <- c("Unadjusted", "U-adjusted", "G:U-adjusted", "G:E-adjusted")

# Rate-vs-parameter curves (power sweeps) and rate bar charts (type I error)
# are the only two figure forms in the simulation, so they get one definition
# each rather than one per panel.
sim_curve <- function(dat, xvar, xlab, ylab = paste0("Rejection rate at alpha = ", SIM_ALPHA)) {
  ggplot(dat, aes({{ xvar }}, rate, color = test, fill = test, shape = test)) +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey50") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2) +
    scale_color_manual(values = sim_cols) +
    scale_fill_manual(values = sim_cols) +
    scale_shape_manual(values = sim_shapes) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
    labs(x = xlab, y = ylab, color = NULL, fill = NULL, shape = NULL) +
    theme(legend.position = "bottom")
}

sim_bars <- function(dat, title, fill = "#1b9e77") {
  pd <- position_dodge(width = 0.7)
  ggplot(dat, aes(model, rate, alpha = condition)) +
    geom_col(fill = fill, color = "grey20", width = 0.7, position = pd) +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.18, position = pd) +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey40") +
    # Each panel uses a different fill, so the alpha key is drawn in a neutral
    # grey: identical guides across panels are what lets patchwork collect them
    # into the single shared legend below.
    scale_alpha_manual(values = c(0.45, 1.0),
                       guide = guide_legend(override.aes = list(fill = "grey30"))) +
    scale_x_discrete(labels = scales::label_wrap(14)) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(x = NULL, y = "Type I error rate of GxM test", alpha = NULL, title = title) +
    theme(axis.text.x = element_text(size = 8))
}

# Shared reshaping: map raw keys onto display labels and order the factors.
label_tests <- function(dat) mutate(dat, test = unname(test_labels[test]))
label_icc   <- function(dat) mutate(dat, condition = factor(unname(icc_labels[icc_condition]),
                                                            levels = unname(icc_labels)))
```

## Causal scenarios

```{r sim-dags, fig.width = 11, fig.asp = 0.30}
# The notebook writes the DAG geometry already resolved (edges trimmed back off
# the node labels, modifier arrows anchored to the midpoint of the edge they
# modify, coefficient labels offset perpendicular to their arrow), so this only
# has to draw it. Rendering here rather than in the notebook keeps every
# manuscript figure on one styling layer.
build_dag <- function(which_scenarios) {
  if (!have(dag_nodes) || !have(dag_edges)) return(NULL)
  e <- filter(dag_edges, scenario %in% which_scenarios)
  n <- filter(dag_nodes, scenario %in% which_scenarios)
  ggplot() +
    geom_segment(data = e,
                 aes(x = x, y = y, xend = xend, yend = yend,
                     linetype = kind, color = kind),
                 arrow = arrow(length = unit(0.10, "in"), type = "closed"),
                 linewidth = 0.55) +
    geom_text(data = e, aes(x = label_x, y = label_y, label = label, color = kind),
              parse = TRUE, size = 3.7, show.legend = FALSE) +
    geom_label(data = n, aes(x = x, y = y, label = node),
               size = 4.6, fontface = "bold", fill = "white",
               label.padding = unit(0.22, "lines")) +
    facet_wrap(~ scenario, nrow = 1, labeller = labeller(scenario = scenario_labels)) +
    scale_linetype_manual(values = c(main = "solid", mod = "dashed")) +
    scale_color_manual(values    = c(main = "grey20", mod = "firebrick")) +
    coord_cartesian(xlim = c(-0.2, 2.2), ylim = c(-0.4, 1.5), clip = "off", expand = FALSE) +
    theme_void(base_size = 13) +
    theme(legend.position = "none",
          strip.text    = element_text(face = "bold"),
          panel.spacing = unit(0.2, "lines"),
          plot.margin   = margin(10, 2, 2, 2))   # Room for the strip title in the combined figure.
}

dag_all <- build_dag(c("A", "B", "C", "D"))
if (!is.null(dag_all)) print(dag_all) else
  cat("results/simulation/dag_nodes.csv and dag_edges.csv not available - skipping.\n")
```

Solid arrows are main effects; dashed red arrows are interaction terms, drawn
landing on the edge they modify. Scenario A is the mediation model the screen
assumes; B, C and D are the non-mediating structures that can still produce a
GxM signal.

## Power under true mediation (Scenario A)

```{r sim-power, fig.asp = 0.5}
if (have(sim_power)) {
  power_panel <<- sim_curve(label_tests(sim_power), beta_GxM,
                            expression(paste("True ", beta[GxM])))
  print(power_panel)

  write_csv(sim_power, "output/manuscript/sim_power.csv")
} else cat("results/simulation/power_summary.csv not available - skipping.\n")
```

The `G:E`-adjusted GxM test is the remedy deployed against the spurious
scenarios below. Comparing the per-replicate GxM z-statistic with and without
that adjustment shows it is free under true mediation: the two statistics fall
on the identity line. Agreement is summarised within each true `beta_GxM`, since
a correlation pooled across effect sizes is inflated by the spread between them.

```{r sim-power-z, fig.asp = 0.6}
if (have(sim_power_z)) {
  z_subtitle <- if (have(sim_z_agree)) {
    sprintf("Median within-effect r = %.3f\nMax |z difference| = %.2f",
            median(sim_z_agree$r), max(sim_z_agree$max_abs_dz))
  } else NULL

  power_z_panel <<-
    ggplot(sim_power_z, aes(z_gxm, z_gxm_adj, color = factor(beta_GxM))) +
      geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
      geom_point(alpha = 0.5, size = 1.3) +
      scale_color_brewer(palette = "Greens", name = expression(paste("True ", beta[GxM]))) +
      coord_equal() +
      labs(x = "GxM z (unadjusted)", y = "GxM z (G:E-adjusted)",
           subtitle = z_subtitle) +
      guides(color = guide_legend(nrow = 1, title.position = "top")) +   # Else clipped in the combined figure.
      theme(legend.position = "bottom")
  print(power_z_panel)
} else cat("results/simulation/power_z_replicates.csv not available - skipping.\n")
```

Rejection rates say nothing about where the estimate sits. Bias is against the
true structural `beta_GxM`, so under measurement error in M it includes the
expected attenuation; coverage is of the nominal 95% interval.

```{r sim-power-bias}
if (have(sim_bias)) {
  sim_bias %>%
    transmute(`True beta_GxM` = beta_GxM, ICC_E, ICC_M,
              Model    = recode(model, gxm = "Unadjusted", gxm_adj = "G:E-adjusted"),
              `Mean estimate (95% CI)` = sprintf("%.4f (%.4f, %.4f)", mean_est, est_lo, est_hi),
              Bias     = round(bias, 4),
              Coverage = sprintf("%.2f (%.2f, %.2f)", coverage, coverage_lo, coverage_hi),
              Replicates = n_used) %>%
    kable(caption = "Bias and 95% interval coverage of the GxM estimate, Scenario A") %>%
    kable_styling(full_width = FALSE)
} else cat("results/simulation/power_bias_summary.csv not available - skipping.\n")
```

## Dilution identity

The conceptual model claims the induced GxE interaction is a diluted image of
the GxM interaction, `beta_GxE ≈ alpha * beta_GxM`, with `alpha` the standardized
E-M association, and the applied dilution check above leans on it. Under
Scenario A (`beta_GxM` fixed, `beta_EM` swept, `ICC_E` and `ICC_M` each 0.5 or 1),
each replicate standardizes Y, E and M and compares the observed GxE estimate with
`alpha_hat * beta_GxM_hat`, using both the marginal GxM estimate the identity
concerns and the G:E-adjusted estimate the applied analysis reports.

```{r sim-dilution, fig.width = 12, fig.asp = 0.75}
if (have(sim_dil_rep) && have(sim_dil)) {
  dil_estimate_labels <- c(marginal = "Marginal GxM", adjusted = "G:E-adjusted GxM")
  label_dilution <- function(dat) {
    dat %>%
      arrange(desc(ICC_E), desc(ICC_M)) %>%
      mutate(condition = fct_inorder(sprintf("ICC_E = %.1f, ICC_M = %.1f", ICC_E, ICC_M)),
             estimate  = factor(unname(dil_estimate_labels[estimate]),
                                levels = unname(dil_estimate_labels)))
  }

  dil_scatter <- sim_dil_rep %>%
    label_dilution() %>%
    ggplot(aes(predicted_gxe, beta_gxe, color = beta_EM)) +
    geom_hline(yintercept = 0, color = "grey85") +
    geom_vline(xintercept = 0, color = "grey85") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.4, size = 0.8) +
    facet_grid(estimate ~ condition) +
    scale_color_viridis_c(name = expression(beta[EM])) +
    labs(x = expression("Predicted " * beta[GxE] == hat(alpha) %*% hat(beta)[GxM]),
         y = expression("Observed " * hat(beta)[GxE]))

  # At beta_EM = 0 the predicted GxE is zero and the ratio has no estimand.
  dil_ratio <- sim_dil %>%
    filter(beta_EM > 0) %>%
    label_dilution() %>%
    mutate(beta_EM_label = fct_inorder(sprintf("beta[EM] == %.2f", beta_EM)),
           ICC_M = factor(ICC_M)) %>%
    ggplot(aes(ICC_E, ratio_median, color = ICC_M, group = ICC_M)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2) +
    facet_grid(estimate ~ beta_EM_label, labeller = labeller(beta_EM_label = label_parsed)) +
    scale_x_continuous(breaks = sort(unique(sim_dil$ICC_E))) +
    scale_color_brewer(palette = "Dark2", name = expression(ICC[M])) +
    labs(x = expression(ICC[E]),
         y = expression("Median " * hat(beta)[GxE] / (hat(alpha) %*% hat(beta)[GxM])))

  print(dil_scatter / dil_ratio + plot_layout(heights = c(1.1, 1)))
} else cat("results/simulation/dilution_replicates.csv and dilution_summary.csv not available - skipping.\n")
```

## Comparison with formal mediated-moderation testing

```{r sim-medmod, fig.asp = 0.5}
if (have(sim_medmod)) {
  # At beta_EM = 0 there is no indirect effect, so the formal test's rejection
  # rate there is type I error on a power axis. GxE and GxM keep that point.
  dat <- sim_medmod %>%
    filter(test != "p_med" | beta_EM > 0) %>%
    label_tests() %>%
    mutate(test      = factor(test, levels = c("GxE", "GxM", "Mediation")),
           condition = factor(unname(medmod_labels[condition]),
                              levels = unname(medmod_labels)))

  print(sim_curve(dat, beta_EM, expression(paste("E-M coupling ", beta[EM]))) +
          facet_wrap(~ condition))
} else cat("results/simulation/medmod_summary.csv not available - skipping.\n")
```

## Sensitivity: power across the full ICC_E x ICC_M grid

```{r sim-icc-grid, fig.asp = 0.95}
if (have(sim_icc)) {
  dat <- sim_icc %>%
    label_tests() %>%
    mutate(ICC_E_label = factor(sprintf("ICC[E] == %.2f", ICC_E)),
           ICC_M_label = factor(sprintf("ICC[M] == %.2f", ICC_M)))

  print(sim_curve(dat, beta_GxM, expression(paste("True ", beta[GxM]))) +
          facet_grid(ICC_E_label ~ ICC_M_label, labeller = label_parsed) +
          scale_x_continuous(breaks = c(0, 0.1, 0.2)))   # Five breaks collide across facets.
} else cat("results/simulation/icc_grid_summary.csv not available - skipping.\n")
```

## Type I error under the non-mediating scenarios

```{r sim-tie, fig.width = 11, fig.asp = 0.42}
tie_variant_panel <- function(dat, title, fill) {
  if (!have(dat)) return(NULL)
  dat %>%
    label_icc() %>%
    mutate(model = factor(unname(variant_labels[variant]),
                          levels = unname(variant_labels))) %>%
    sim_bars(title, fill)
}

panel_b <<- tie_variant_panel(sim_tie_b, "Upstream GxE", "#7570b3")
panel_c <<- tie_variant_panel(sim_tie_c, "Reverse causation", "#d95f02")

panel_d <<- if (have(sim_tie)) {
  sim_tie %>%
    filter(scenario == "D", test != "p_gxe") %>%
    label_icc() %>%
    label_tests() %>%
    mutate(model = factor(test, levels = D_MODELS)) %>%
    sim_bars("E-M confounder", "#1b9e77")
} else NULL

panels <- compact(list(panel_b, panel_c, panel_d))
if (length(panels)) {
  # Scenario D has four bars against three in B and C, so it gets the extra width.
  widths <- ifelse(map_lgl(panels, ~ identical(.x, panel_d)), 1.3, 1)
  print(wrap_plots(panels, nrow = 1) +
          plot_layout(guides = "collect", widths = widths) &
          theme(legend.position = "bottom"))
} else cat("No scenario B/C/D summaries available - skipping.\n")
```

```{r sim-tie-table}
if (have(sim_tie)) {
  # An export from before the notebook kept GxM estimates has no estimate columns.
  if (!"mean_est" %in% names(sim_tie))
    sim_tie <- mutate(sim_tie, mean_est = NA_real_, est_lo = NA_real_, est_hi = NA_real_)

  sim_tie %>%
    filter(!is.na(rate)) %>%
    label_tests() %>%
    transmute(Scenario = scenario,
              Measurement = unname(icc_labels[icc_condition]),
              Model = test,
              `Rejection rate` = round(rate, 3),
              `95% CI` = sprintf("%.3f - %.3f", lo, hi),
              # Blank for the GxE row, which estimates no GxM coefficient.
              `Mean GxM estimate (95% CI)` = ifelse(is.na(mean_est), "",
                sprintf("%.4f (%.4f, %.4f)", mean_est, est_lo, est_hi)),
              Replicates = n_used) %>%
    arrange(Scenario, Measurement, Model) %>%
    kable(caption = "Type I error of the GxM test by scenario and adjustment") %>%
    kable_styling(full_width = FALSE)

  write_csv(sim_tie, "output/manuscript/sim_type1_error.csv")
}
```

The inflation that survives the `G:E` adjustment in Scenario B is a bias rather
than an inflated variance, so the mean spurious coefficient stays put as N grows
while the rejection rate climbs.

```{r sim-nsweep}
if (have(sim_nsweep)) {
  sim_nsweep %>%
    transmute(N = scales::comma(N),
              Model = recode(model, unadj = "Unadjusted", adj_GxE = "G:E-adjusted"),
              `Rejection rate (95% CI)` = sprintf("%.3f (%.3f, %.3f)", rate, lo, hi),
              `Mean GxM estimate (95% CI)` = sprintf("%.4f (%.4f, %.4f)", mean_est, est_lo, est_hi),
              Replicates = n_used) %>%
    kable(caption = "Scenario B, realistic measurement: rejection rate and mean GxM estimate by sample size") %>%
    kable_styling(full_width = FALSE)
} else cat("results/simulation/tie_scenarioB_nsweep_summary.csv not available - skipping.\n")
```

## Combined manuscript figure

```{r sim-combined, fig.width = 12, fig.asp = 0.92}
# Reassembles the panels built above; nothing is recomputed. Missing inputs drop
# their row rather than failing the figure.
top_row <- if (!is.null(build_dag("A")) && exists("power_panel") && exists("power_z_panel")) {
  build_dag("A") + power_panel + power_z_panel + plot_layout(widths = c(1, 1.4, 1.2))
} else NULL

mid_row <- build_dag(c("B", "C", "D"))

tie_panels <- compact(list(
  if (exists("panel_b")) panel_b, if (exists("panel_c")) panel_c,
  if (exists("panel_d")) panel_d))
bot_row <- if (length(tie_panels)) {
  wrap_plots(tie_panels, nrow = 1) + plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
} else NULL

rows <- compact(list(top_row, mid_row, bot_row))
if (length(rows) == 3) {
  combined_fig <- rows[[1]] / rows[[2]] / rows[[3]] +
    plot_layout(heights = c(1.0, 0.8, 1.3)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 16))
  print(combined_fig)
} else {
  cat("Need the DAG tables, the power summaries and the type I error summaries",
      "to assemble the combined figure -", length(rows), "of 3 rows available.\n")
}
```

## Sensitivity: nonnegative (gamma-distributed) exposure

Rejection rates under a Gamma(shape = 4) exposure against the normally
distributed exposure, under perfect and realistic measurement. The gamma
exposure is scaled to the variance of E in each normal-E scenario and measured
with error through a Gaussian copula that keeps it nonnegative, with the
confounder U in Scenario D entering through the same copula, so the two columns
differ only in the shape of E.

```{r sim-gamma, fig.asp = 0.8}
if (have(sim_gamma) && have(sim_tie)) {
  # An export from before the gamma arm gained measurement error ran at perfect
  # measurement only, so it compares against that condition alone.
  if (!"icc_condition" %in% names(sim_gamma)) sim_gamma$icc_condition <- "perfect"

  normal_arm <- sim_tie %>%
    filter(icc_condition %in% sim_gamma$icc_condition) %>%
    transmute(scenario, icc_condition, model = unname(test_labels[test]),
              rate, e_dist = "Normal E")

  gamma_arm <- sim_gamma %>%
    transmute(scenario, icc_condition, model = unname(test_labels[test]),
              rate, e_dist = "Gamma E (shape 4)")

  bind_rows(normal_arm, gamma_arm) %>%
    filter(!is.na(rate), model %in% D_MODELS) %>%
    label_icc() %>%
    mutate(model    = factor(model, levels = D_MODELS),
           e_dist   = factor(e_dist, levels = c("Normal E", "Gamma E (shape 4)")),
           scenario = paste("Scenario", scenario)) %>%
    ggplot(aes(model, rate, fill = e_dist)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65, color = "grey20") +
    geom_hline(yintercept = SIM_ALPHA, linetype = "dashed", color = "grey40") +
    facet_grid(condition ~ scenario, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = c("Normal E" = "#7570b3", "Gamma E (shape 4)" = "#d95f02")) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(x = NULL, y = "Type I error rate of GxM test", fill = "Exposure distribution") +
    theme(legend.position = "bottom", axis.text.x = element_text(size = 8, angle = 30, hjust = 1))
} else cat("Need tie_gamma_summary.csv and tie_summary.csv - skipping.\n")
```

# Session

```{r session-info}
sessionInfo()
```
